---
title: "Model Diagnostics and Visualization"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates model diagnostics and visualization using **mgcv** and **gratia** in R. We fit the same Gu & Wahba (1991) example as the Julia vignette and compare diagnostic outputs.

## Setup

```{r}
library(mgcv)
library(gratia)
```

## Load Gu–Wahba example data

```{r}
dat <- read.csv("../data.csv")
n <- nrow(dat)
x0 <- dat$x0; x1 <- dat$x1; x2 <- dat$x2; x3 <- dat$x3
y <- dat$y

f0 <- function(x) 2 * sin(pi * x)
f1 <- function(x) exp(2 * x)
f2 <- function(x) 0.2 * x^11 * (10 * (1 - x))^6 + 10 * (10 * x)^3 * (1 - x)^10
f3 <- function(x) rep(0, length(x))

head(dat)
```

## Fit the GAM

```{r}
m <- gam(y ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
             s(x2, k = 10, bs = "cr") + s(x3, k = 10, bs = "cr"),
         data = dat, method = "REML")
summary(m)
```

## Model overview

gratia's `overview` provides a tidy summary of all smooth terms.

```{r}
overview(m)
```

## Smooth estimates

`smooth_estimates` evaluates each smooth on a regular grid with pointwise standard errors.

```{r}
se <- smooth_estimates(m, n = 200)
head(se)
```

### Plotting smooth estimates with true functions

```{r}
#| fig-width: 8
#| fig-height: 8
true_fns <- list(f0, f1, f2, f3)
x_vars <- c("x0", "x1", "x2", "x3")
fn_labels <- c(
  expression(f[0](x[0]) == 2*sin(pi*x)),
  expression(f[1](x[1]) == exp(2*x)),
  expression(f[2](x[2]) ~ "Gu-Wahba"),
  expression(f[3](x[3]) == 0)
)

par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (i in 1:4) {
  se_i <- smooth_estimates(m, select = paste0("s(", x_vars[i], ")"), n = 200)
  x_grid <- se_i[[x_vars[i]]]
  est <- se_i[[".estimate"]]
  se_vals <- se_i[[".se"]]

  f_grid <- true_fns[[i]](x_grid)
  f_grid <- f_grid - mean(f_grid)

  ylims <- range(c(est - 2 * se_vals, est + 2 * se_vals, f_grid))
  plot(x_grid, est, type = "l", lwd = 2, col = "steelblue",
       ylim = ylims, xlab = x_vars[i], ylab = paste0("f(", x_vars[i], ")"),
       main = fn_labels[i])
  polygon(c(x_grid, rev(x_grid)),
          c(est - 2 * se_vals, rev(est + 2 * se_vals)),
          col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)
  lines(x_grid, f_grid, lty = 2, lwd = 2, col = "red")
  legend("topright", c("estimate", "truth"), lty = c(1, 2),
         col = c("steelblue", "red"), lwd = 2, cex = 0.7, bg = "white")
}
```

### gratia's built-in `draw`

```{r}
#| fig-width: 8
#| fig-height: 6
draw(m, residuals = TRUE, rug = FALSE)
```

## Partial residuals

Partial residuals overlaid on smooth estimates show how well the smooth captures the data pattern.

```{r}
#| fig-width: 8
#| fig-height: 8
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
x_data <- list(x0, x1, x2, x3)
pr <- partial_residuals(m)

for (i in 1:4) {
  se_i <- smooth_estimates(m, select = paste0("s(", x_vars[i], ")"), n = 200)
  x_grid <- se_i[[x_vars[i]]]
  est <- se_i[[".estimate"]]
  se_vals <- se_i[[".se"]]

  plot(x_grid, est, type = "l", lwd = 2, col = "steelblue",
       xlab = x_vars[i], ylab = paste0("f(", x_vars[i], ")"),
       main = paste0("s(", x_vars[i], ")"),
       ylim = range(c(est - 2 * se_vals, est + 2 * se_vals, pr[[i]])))
  polygon(c(x_grid, rev(x_grid)),
          c(est - 2 * se_vals, rev(est + 2 * se_vals)),
          col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)
  points(x_data[[i]], pr[[i]], pch = 16, cex = 0.5, col = adjustcolor("grey40", 0.4))
}
```

## Derivatives

`derivatives` computes finite-difference derivatives with confidence intervals.

```{r}
d <- derivatives(m, type = "central", n = 200)
head(d)
```

### Plotting derivatives

```{r}
#| fig-width: 8
#| fig-height: 8
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (i in 1:4) {
  d_i <- derivatives(m, select = paste0("s(", x_vars[i], ")"),
                     type = "central", n = 200)
  x_grid <- d_i[[x_vars[i]]]
  deriv <- d_i[[".derivative"]]
  lower <- d_i[[".lower_ci"]]
  upper <- d_i[[".upper_ci"]]
  sig <- (lower > 0) | (upper < 0)

  ylims <- range(c(lower, upper))
  plot(x_grid, deriv, type = "l", lwd = 2, col = "steelblue",
       ylim = ylims, xlab = x_vars[i], ylab = "derivative",
       main = paste0("d/d", x_vars[i]))
  polygon(c(x_grid, rev(x_grid)),
          c(lower, rev(upper)),
          col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)
  abline(h = 0, lty = 2, col = "grey50")

  if (any(sig)) {
    rug(x_grid[sig], col = "red", lwd = 2, side = 1)
  }
}
```

Second-order derivatives:

```{r}
#| fig-width: 6
#| fig-height: 4
d2 <- derivatives(m, select = "s(x0)", order = 2, n = 200)
x_grid <- d2[["x0"]]
deriv <- d2[[".derivative"]]
lower <- d2[[".lower_ci"]]
upper <- d2[[".upper_ci"]]

plot(x_grid, deriv, type = "l", lwd = 2, col = "darkorange",
     ylim = range(c(lower, upper)),
     xlab = expression(x[0]), ylab = "second derivative",
     main = "Second derivative of s(x0)")
polygon(c(x_grid, rev(x_grid)),
        c(lower, rev(upper)),
        col = adjustcolor("darkorange", alpha.f = 0.2), border = NA)
abline(h = 0, lty = 2, col = "grey50")
```

## Model diagnostics with `appraise`

`appraise` produces the four standard diagnostic plots: QQ, residuals vs linear predictor, histogram, and observed vs fitted.

```{r}
#| fig-width: 8
#| fig-height: 8
appraise(m)
```

## Posterior simulation

### Smooth samples (spaghetti plot)

`smooth_samples` draws posterior realizations of individual smooth functions.

```{r}
#| fig-width: 8
#| fig-height: 5
ss <- smooth_samples(m, select = "s(x0)", n = 50, seed = 42)

se_x0 <- smooth_estimates(m, select = "s(x0)", n = 100)
x_grid <- se_x0[["x0"]]
est <- se_x0[[".estimate"]]

draws <- unique(ss[[".draw"]])
ylims <- range(ss[[".value"]])

plot(NULL, xlim = range(x_grid), ylim = ylims,
     xlab = expression(x[0]), ylab = expression(f(x[0])),
     main = "50 posterior draws of s(x0)")
for (dr in draws) {
  idx <- ss[[".draw"]] == dr
  lines(ss[["x0"]][idx], ss[[".value"]][idx],
        col = adjustcolor("steelblue", 0.15), lwd = 0.5)
}
lines(x_grid, est, col = "red", lwd = 2)
legend("topright", c("posterior draws", "point estimate"),
       col = c("steelblue", "red"), lwd = c(1, 2), cex = 0.8)
```

### Fitted samples

`fitted_samples` draws posterior samples of fitted values.

```{r}
fs <- fitted_samples(m, n = 200, seed = 42)
cat("Fitted samples: ", nrow(fs), "rows x", length(unique(fs[[".draw"]])), "draws\n")
```

## Concurvity

```{r}
model_concurvity(m)
```

Pairwise concurvity:

```{r}
model_concurvity(m, pairwise = TRUE)
```

## Basis dimension check

```{r}
k.check(m)
```

## Uncertainty when the smoothing parameters are estimated

mgcv carries the same three degrees-of-freedom measures as GAM.jl, in
`m$edf`, `m$edf1` and `m$edf2`, and the same corrected covariance in `m$Vc`.
These are the quantities behind `predict.gam(..., unconditional = TRUE)`.

```{r}
cat(sprintf("edf_total  : %.4f\n", sum(m$edf)))
cat(sprintf("sum(edf1)  : %.4f\n", sum(m$edf1)))
cat(sprintf("sum(edf2)  : %.4f\n", sum(m$edf2)))
cat(sprintf("per-smooth edf: %s\n",
    paste(round(sapply(m$smooth,
        function(s) sum(m$edf[s$first.para:s$last.para])), 3), collapse = ", ")))
```

```{r}
Vp <- m$Vp; Vc <- m$Vc
r <- sqrt(diag(Vc)) / sqrt(diag(Vp))
cat(sprintf("rel. Frobenius |Vc - Vp|/|Vp| : %.4f\n",
    norm(Vc - Vp, "F") / norm(Vp, "F")))
cat(sprintf("SE ratio Vc/Vp - min %.4f, median %.4f, max %.4f\n",
    min(r), median(r), max(r)))
cat(sprintf("intercept SE ratio : %.4f\n", r[1]))
cat(sprintf("REML score (gcv.ubre) : %.4f\n", m$gcv.ubre))
cat(sprintf("AIC : %.4f\n", AIC(m)))
```

These agree closely with the Julia vignette, which is the point of the
comparison — the `Vc` machinery is a faithful port, not a re-derivation:

| Quantity | GAM.jl | mgcv 1.9-4 |
|----------|--------|------------|
| `edf_total` | 17.3502 | 17.3458 |
| `sum(edf1)` | 20.2004 | 20.1950 |
| `sum(edf2)` | 19.6689 | 19.6644 |
| $\lVert V_c - V_p\rVert/\lVert V_p\rVert$ | 0.2794 | 0.2794 |
| max SE ratio | 1.4279 | 1.4285 |
| REML score | 881.9628 | 881.9628 |
| AIC | 1752.8195 | 1752.8194 |

The residual differences (in the fifth significant figure) come from the two
implementations converging to very slightly different smoothing parameters,
not from a difference in the formulas.

Interval widths with and without the correction:

```{r}
p_con <- predict(m, se.fit = TRUE, unconditional = FALSE)
p_unc <- predict(m, se.fit = TRUE, unconditional = TRUE)
cat(sprintf("mean SE conditional (Vp)   : %.5f\n", mean(p_con$se.fit)))
cat(sprintf("mean SE unconditional (Vc) : %.5f  (x%.4f)\n",
    mean(p_unc$se.fit), mean(p_unc$se.fit) / mean(p_con$se.fit)))
```

## Visualizing 2-D smooths

`vis.gam` is mgcv's 2-D surface plot; the Julia vignette's `vis_gam` returns
the same surface as data (`x1`, `x2`, `z`) for plotting with any backend.
Both accept `too.far` / `too_far` to mask grid cells far from the data.

```{r}
d2 <- read.csv("../data_2d.csv")
m2d <- gam(y ~ s(x, z, k = 40, bs = "tp"), data = d2, method = "REML")
cat(sprintf("2-D fit: edf = %.3f\n", sum(m2d$edf)))
```

```{r}
#| fig-width: 9
#| fig-height: 4.5
op <- par(mfrow = c(1, 2))
vis.gam(m2d, view = c("x", "z"), plot.type = "contour",
        too.far = 0.1, main = "vis.gam contour")
vis.gam(m2d, view = c("x", "z"), plot.type = "persp",
        theta = 30, phi = 25, main = "vis.gam perspective")
par(op)
```

## Summary

Both GAM.jl and gratia provide equivalent diagnostic interfaces:

| gratia / mgcv (R) | GAM.jl (Julia) |
|------------|-----------------|
| `overview()` | `overview()` |
| `smooth_estimates()` | `smooth_estimates()` |
| `derivatives()` | `derivatives()` |
| `appraise()` | `appraise()` |
| `posterior_samples()` | `posterior_samples()` |
| `fitted_samples()` | `fitted_samples()` |
| `smooth_samples()` | `smooth_samples()` |
| `partial_residuals()` | `partial_residuals()` |
| `model_concurvity()` | `concurvity()` |
| `k.check()` | `k_check()` |
| `m$edf`, `m$edf1`, `m$edf2` | `edf()`, `m.edf1`, `edf2()` |
| `m$Vc` | `vcov_corrected()` |
| `unconditional = TRUE` | `unconditional = true` |
| `m$gcv.ubre` | `sp_criterion()` |
| `vis.gam()` | `vis_gam()`, `gamcontour()` |
