---
title: "Model Selection and Diagnostics: R Comparison"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This companion runs the **same** seven-step model-selection workflow on the
**same** `data.csv` as `14_model_selection.qmd`, using mgcv, so the two
printouts can be compared step by step:

1. fit, and ask whether each basis is **large enough** (`k.check`)
2. ask whether any term is **unnecessary** (`select = TRUE`)
3. ask whether the terms are **mutually identifiable** (`concurvity`)
4. compare **families** by AIC
5. check the **distributional assumptions** (`gam.check`, `qq.gam`)
6. find **influential observations** (leverage, Cook's distance)
7. look at **partial residuals** to see what the smooths are fitting

```{r}
library(mgcv)
```

## The data

The Gu & Wahba four-term example with three deliberate complications:

- $f_0(x) = 2\sin(\pi x)$, $f_1(x) = e^{2x}$,
  $f_2(x) = 0.2x^{11}(10(1-x))^6 + 10(10x)^3(1-x)^{10}$
- $f_3(x) = 0$ — a **null smooth**, in the model but absent from the truth
- `x4` is `x1` plus small noise — a **near-duplicate covariate**, creating concurvity
- observation 100 has **+15 added** — a gross outlier

```{r}
df <- read.csv("../data.csv")
cat(sprintf("n = %d, columns: %s\n", nrow(df), paste(names(df), collapse = ", ")))
head(df, 4)
```

## Step 1 — Fit, then check basis dimensions

```{r}
m <- gam(y ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
             s(x2, k = 10, bs = "cr") + s(x3, k = 10, bs = "cr"),
         data = df, method = "REML")
summary(m)
```

`k.check` reports the basis dimension, the effective degrees of freedom, the
k-index and a permutation p-value per smooth. A **low k-index with a small
p-value means the basis is too small**.

```{r}
k.check(m)
```

All four p-values are comfortable and the edf sit well below `k`, so the bases
are large enough — the same conclusion the Julia vignette reaches with
`k_check(m)`.

## Step 2 — Is any term unnecessary?

`select = TRUE` adds a penalty on each smooth's null space, so a term with no
support can be shrunk to *zero* effective degrees of freedom, not merely to a
straight line.

```{r}
ms <- gam(y ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
              s(x2, k = 10, bs = "cr") + s(x3, k = 10, bs = "cr"),
          data = df, method = "REML", select = TRUE)

cmp <- data.frame(
  term        = c("s(x0)", "s(x1)", "s(x2)", "s(x3)"),
  edf_plain   = round(summary(m)$edf, 3),
  edf_select  = round(summary(ms)$edf, 3)
)
print(cmp, row.names = FALSE)
```

`s(x3)` — the null smooth — collapses to essentially zero: the term is removed
from the model. The genuine terms keep their structure, though the extra
null-space penalty shrinks them a little too. That is the trade term selection
makes, and it is why `select = TRUE` is a modelling choice rather than a free
lunch.

```{r}
cat(sprintf("deviance explained: plain %.3f, select %.3f\n",
            summary(m)$dev.expl, summary(ms)$dev.expl))
```

## Step 3 — Concurvity

Concurvity is the smooth analogue of collinearity: how well can one smooth be
reproduced by the others? `concurvity(m, full = TRUE)` returns mgcv's three
measures — `worst` is the pessimistic bound, `observed` uses the fitted values,
`estimate` is a squared-Frobenius ratio. Values near 1 are trouble.

```{r}
round(concurvity(m, full = TRUE), 3)
```

Low across the board — the covariates are independent draws. Now swap in `x4`,
which is `x1` plus small noise:

```{r}
m_conc <- gam(y ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
                  s(x2, k = 10, bs = "cr") + s(x4, k = 10, bs = "cr"),
              data = df, method = "REML")
round(concurvity(m_conc, full = TRUE), 3)
```

`s(x1)` and `s(x4)` now flag high concurvity, exactly as they should: the model
cannot tell their contributions apart. The remedy is to drop one, not to reach
for a bigger basis.

## Step 4 — Comparing families by AIC

```{r}
m_gauss <- gam(y ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
                   s(x2, k = 10, bs = "cr"), data = df, method = "REML")

df_pos <- df
df_pos$ypos <- df$y - min(df$y) + 0.5          # shift to positive support
m_gamma <- gam(ypos ~ s(x0, k = 10, bs = "cr") + s(x1, k = 10, bs = "cr") +
                      s(x2, k = 10, bs = "cr"), data = df_pos,
               family = Gamma(link = "log"), method = "REML")

cat(sprintf("Gaussian  AIC = %9.2f  (edf %.2f)\n", AIC(m_gauss), sum(m_gauss$edf)))
cat(sprintf("Gamma/log AIC = %9.2f  (edf %.2f)\n", AIC(m_gamma), sum(m_gamma$edf)))
```

The same caveat as in the Julia vignette applies: these two models have
**different responses** (`y` versus a shifted `ypos`), so their AIC values are
*not* comparable with each other. They are shown to illustrate the call. AIC
comparisons are only meaningful across models of the same response.

## Step 5 — Distributional assumptions

`gam.check` produces the four standard residual plots and reprints the basis
check. mgcv's `qq.gam` uses a simulated reference by default, matching the
`method = :simulate` default of GAM.jl's `appraise`.

```{r fig.width=8, fig.height=7}
par(mfrow = c(2, 2))
gam.check(m)
```

The outlier at observation 100 is visible in the QQ tail and in the
residuals-versus-linear-predictor panel.

## Step 6 — Influential observations

The hat-matrix diagonal sums to the model's effective degrees of freedom, a
useful sanity check. Cook's distance combines leverage with residual size.

```{r}
lev <- influence(m)                       # hat diagonal for a fitted gam
# mgcv's m$edf is per-coefficient and already includes the intercept, so its
# sum is the model's total effective degrees of freedom.
edf_total <- sum(m$edf)
cat(sprintf("sum(leverage) = %.3f   total edf = %.3f   (should match)\n",
            sum(lev), edf_total))

r    <- residuals(m, type = "response")
phi  <- m$sig2
cook <- (r^2 * lev) / (phi * edf_total * (1 - lev)^2)
cat(sprintf("largest Cook's distance at observation %d (value %.4f, median %.5f)\n",
            which.max(cook), max(cook), median(cook)))
```

The planted outlier is identified as the most influential point.

```{r fig.width=8, fig.height=3.5}
par(mfrow = c(1, 2))
plot(lev, pch = 16, cex = 0.5, col = "steelblue",
     xlab = "Observation", ylab = "Leverage", main = "Leverage")
plot(cook, pch = 16, cex = 0.5, col = "steelblue",
     xlab = "Observation", ylab = "Cook's distance", main = "Influence")
points(which.max(cook), max(cook), col = "red", pch = 16, cex = 1.2)
```

## Step 7 — Partial residuals

Partial residuals are the working residuals plus the term's own contribution.
Overlaying them on the estimated smooth shows whether the fit follows the data
or is being dragged by a few points. `plot.gam(..., residuals = TRUE)` does
this directly.

```{r fig.width=8, fig.height=7}
par(mfrow = c(2, 2))
plot(m, residuals = TRUE, pch = 16, cex = 0.4, shade = TRUE, seWithMean = TRUE)
```

`s(x3)` is visibly flat with a band covering zero — the graphical counterpart of
the `select = TRUE` result in step 2.

## Summary

| Question | mgcv | GAM.jl |
|---|---|---|
| Basis big enough? | `k.check(m)` | `k_check(m)` |
| Term needed? | `select = TRUE` | `select = true` |
| Terms identifiable? | `concurvity(m, full = TRUE)` | `concurvity(m; full = true)` |
| Right family? | `AIC(m)` | `aic(m)` |
| Assumptions OK? | `gam.check(m)`, `qq.gam` | `appraise(m)` |
| Any point dominating? | `influence(m)`, Cook's distance | `leverage(m)`, `cooksdistance(m)` |
| What is each smooth fitting? | `plot(m, residuals = TRUE)` | `partial_residuals(m)` |

### Numerical agreement with the Julia vignette

Run side by side on this dataset, the two packages agree closely:

| Quantity | GAM.jl | mgcv |
|---|---|---|
| edf `s(x0)` | 3.202 | 3.191 |
| edf `s(x1)` | 2.567 | 2.566 |
| edf `s(x2)` | 7.798 | 7.798 |
| edf `s(x3)` (null term) | 1.004 | 1.001 |
| edf `s(x3)` under `select` | 0.004 | 0.001 |
| deviance explained | 0.723 | 0.723 |
| concurvity `worst`, `s(x1)` | 0.176 | 0.176 |
| concurvity `worst`, `s(x1)` with `x4` | 0.993 | 0.993 |
| concurvity `worst`, `s(x4)` | 0.993 | 0.993 |
| total edf = Σ leverage | 15.570 | 15.556 |
| most influential observation | 100 | 100 |
| largest Cook's distance | 0.1427 | 0.1427 |
| median Cook's distance | 0.00089 | 0.00089 |

Concurvity agrees to three decimal places, effective degrees of freedom to
about 0.01, and Cook's distances agree to the four decimals printed. Both
packages identify the planted outlier at observation 100, and the null smooth
collapses to essentially zero edf under term selection in both.

### Notes on comparability

- **Term selection** behaves the same way in both packages: the null smooth
  `s(x3)` shrinks to essentially zero edf under `select`, while the genuine
  terms lose a little flexibility to the extra null-space penalty.
- **Concurvity** uses the same three measures and the same definitions, so the
  numbers are directly comparable — as the table shows.
- **Cook's distance** is computed explicitly above because mgcv exposes the hat
  diagonal through `influence()` rather than providing a `cooks.distance`
  method for `gam` objects. Scaling by `sum(m$edf)` — which already counts the
  intercept — reproduces GAM.jl's `cooksdistance(m)` exactly.
- **Smooth-term p-values** can differ slightly: GAM.jl uses a simplification of
  mgcv's `testStat`, so the printed statistic may not match `summary.gam`'s even
  where the conclusion is identical. Empirical rejection rates agree with
  mgcv's to within Monte Carlo error, so test *size* is unaffected.
- **`aic`** in GAM.jl follows mgcv's `AIC()` convention (`edf2`-based) and
  agrees to about 1e-4 on REML fits.
