---
title: "GAMM: R Comparison"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This vignette shows the R equivalent of the GAMM models fit in `10_gamm.qmd`, using `mgcv::gamm()` which internally delegates to `nlme::lme()` for Gaussian models and penalized quasi-likelihood (PQL) for non-Gaussian models.

## Setup

```{r}
library(mgcv)
library(nlme)
```

## Example 1: Gaussian GAMM with Random Intercepts

```{r}
dat <- read.csv("../data_gaussian_gamm.csv")
cat(sprintf("n = %d, subjects = %d\n", nrow(dat), length(unique(dat$subject))))
cat(sprintf("y range: [%.2f, %.2f]\n", min(dat$y), max(dat$y)))
```

### Fitting with `gamm()`

In mgcv, random effects are specified via the `random=` argument, not in the formula:

```{r}
dat$subject <- factor(dat$subject)
m <- gamm(y ~ s(x, k = 15), random = list(subject = ~1), data = dat)
summary(m$gam)
```

### Random effects

```{r}
re <- ranef(m$lme)$subject
cat("Random intercept estimates:\n")
for (lev in rownames(re)) {
  cat(sprintf("  Subject %2s: b = %+.3f\n", lev, re[lev, 1]))
}
```

### Variance components

```{r}
vc <- VarCorr(m$lme)
print(vc)
```

```{r}
sigma_re <- as.numeric(vc[rownames(vc) == "(Intercept)", "StdDev"])
sigma_res <- as.numeric(vc[rownames(vc) == "Residual", "StdDev"])
cat(sprintf("RE σ: %.4f (true: 0.6)\n", sigma_re))
cat(sprintf("Residual σ: %.4f (true: 0.4)\n", sigma_res))
```

### Comparison with true values

```{r}
true_re <- tapply(dat$re_true, dat$subject, function(x) x[1])
cat(sprintf("Correlation of estimated vs true RE: %.4f\n",
            cor(re[, 1], true_re)))
```

### Equivalence with `s(subject, bs="re")`

In mgcv, `s(subject, bs="re")` is the smooth-term equivalent of a random intercept:

```{r}
m_gam <- gam(y ~ s(x, k = 15) + s(subject, bs = "re"),
             data = dat, method = "REML")
cat(sprintf("Fitted values correlation (gamm vs gam+re): %.6f\n",
            cor(fitted(m$lme), fitted(m_gam))))
cat(sprintf("Scale (gamm): %.6f\n", m$gam$scale))
cat(sprintf("Scale (gam):  %.6f\n", m_gam$scale))
```

### Plot: Smooth Effect

```{r}
par(mfrow = c(1, 2))

# Smooth effect
plot(m$gam, shade = TRUE, main = "Population smooth f(x)",
     xlab = "x", ylab = "f(x)")

# Random effects
re_est <- re[, 1]
names(re_est) <- rownames(re)
barplot(sort(re_est), horiz = TRUE, las = 1,
        main = "Random intercepts", xlab = "Estimated b_j",
        col = "steelblue")
abline(v = 0, lty = 2)
```

## Example 2: Poisson GAMM for Count Data

```{r}
dat2 <- read.csv("../data_poisson_gamm.csv")
dat2$site <- factor(dat2$site)
cat(sprintf("n = %d, sites = %d\n", nrow(dat2), length(unique(dat2$site))))
```

### Fitting

For non-Gaussian families, `gamm()` uses PQL (penalized quasi-likelihood):

```{r}
m2 <- gamm(y ~ s(x, k = 15), random = list(site = ~1),
           family = poisson(), data = dat2)
summary(m2$gam)
```

### Random effects

```{r}
re2 <- ranef(m2$lme)$site
true_re2 <- tapply(dat2$re_true, dat2$site, function(x) x[1])
cat(sprintf("RE correlation with truth: %.4f\n",
            cor(re2[, 1], true_re2)))
```

### Plot

```{r}
par(mfrow = c(1, 2))
plot(m2$gam, shade = TRUE, main = "Poisson GAMM: f(x)",
     xlab = "x", ylab = "f(x)")

re2_est <- re2[, 1]
names(re2_est) <- rownames(re2)
barplot(sort(re2_est), horiz = TRUE, las = 1,
        main = "Site random intercepts", xlab = "Estimated b_j",
        col = "darkorange")
abline(v = 0, lty = 2)
```

## Syntax Comparison

| Feature | Julia (GAM.jl) | R (mgcv) |
|---------|----------------|----------|
| Random intercept | `gamm(@gamm_formula(y ~ s(x) + (1\|g)), df)` | `gamm(y ~ s(x), random=list(g=~1), data=df)` |
| `@formula` path | `gamm(@formula(y ~ cr(x, 10) + (1\|g)), df)` | — |
| `re()` shorthand | `gamm(@formula(y ~ cr(x, 10) + re(g)), df)` | — |
| Equivalent GAM | `gam(@gam_formula(y ~ s(x) + s(g, bs=:re)), df)` | `gam(y ~ s(x) + s(g, bs="re"), data=df)` |
| Non-Gaussian | `gamm(..., Poisson())` | `gamm(..., family=poisson())` |
| Extract RE | `ranef(m)` | `ranef(m$lme)` |
| Variance components | `VarCorr(m)` | `VarCorr(m$lme)` |
| Prediction | `predict(m, newdata)` | `predict(m$gam, newdata)` |

### Key differences

1. **Formula syntax**: Julia puts random effects *in the formula* with `(1|group)`, matching lme4/MixedModels.jl conventions. R mgcv uses a separate `random=` argument.

2. **Fitting algorithm**: GAM.jl treats random effects as smooth terms with identity penalty (via PIRLS+REML). R's `gamm()` converts smooths to mixed-model form and delegates to `nlme::lme()` (Gaussian) or uses PQL (non-Gaussian).

3. **Result structure**: Julia returns a `GammModel` with direct access. R returns a list with `$gam` (the GAM component) and `$lme` (the mixed model component).

4. **Interface**: Julia `gamm()` returns a single model object. R requires navigating `m$gam` vs `m$lme` for different aspects of the fit.
