---
title: "Multiple Smooth Terms"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

One of the key strengths of GAMs is the ability to include **multiple smooth terms** additively. The model:

$$
y_i = \beta_0 + f_0(x_{0i}) + f_1(x_{1i}) + f_2(x_{2i}) + f_3(x_{3i}) + \varepsilon_i
$$

estimates each smooth function $f_j$ simultaneously, with automatic smoothness selection for each term.

This vignette uses the classic **Gu–Wahba** simulation (equivalent to `gamSim(eg = 1)` in mgcv) with four smooth functions of varying complexity.

## Setup

```{r}
library(mgcv)
library(gratia)
library(ggplot2)
```

## Loading Gu–Wahba data

The four test functions are:

- $f_0(x) = 2\sin(\pi x)$ — a smooth sinusoid
- $f_1(x) = e^{2x}$ — an exponential
- $f_2(x) = 0.2 x^{11} (10(1-x))^6 + 10(10x)^3 (1-x)^{10}$ — a bumpy function
- $f_3(x) = 0$ — a null smooth (no effect)

```{r}
df <- read.csv("../data.csv")
n <- nrow(df)
x0 <- df$x0; x1 <- df$x1; x2 <- df$x2; x3 <- df$x3
y <- df$y

f0 <- function(x) 2 * sin(pi * x)
f1 <- function(x) exp(2 * x)
f2 <- function(x) 0.2 * x^11 * (10 * (1 - x))^6 + 10 * (10 * x)^3 * (1 - x)^10
f3 <- function(x) rep(0, length(x))

# Center the true functions for identifiability
f0_vals <- f0(x0); f0_vals <- f0_vals - mean(f0_vals)
f1_vals <- f1(x1); f1_vals <- f1_vals - mean(f1_vals)
f2_vals <- f2(x2); f2_vals <- f2_vals - mean(f2_vals)
f3_vals <- f3(x3)

head(df)
```

## Fitting the model

```{r}
m <- gam(y ~ s(x0) + s(x1) + s(x2) + s(x3), data = df, method = "REML")
summary(m)
```

## Model overview

The `gratia::overview()` function summarizes all smooth terms:

```{r}
overview(m)
```

Per-smooth EDF:

```{r}
cat("Per-smooth EDF:\n")
for (i in seq_along(m$smooth)) {
  cat(sprintf("  %-12s EDF = %.2f\n", m$smooth[[i]]$label, pen.edf(m)[i]))
}
cat(sprintf("\nTotal model EDF: %.2f\n", sum(m$edf)))
cat(sprintf("Deviance explained: %.1f%%\n", summary(m)$dev.expl * 100))
```

## Plotting each smooth

We plot each smooth estimate alongside the corresponding true function:

```{r}
#| fig-height: 8
#| fig-width: 8
true_fns <- list(f0, f1, f2, f3)
x_vars <- c("x0", "x1", "x2", "x3")
x_data <- list(x0, x1, x2, x3)
f_centered <- list(f0_vals, f1_vals, f2_vals, f3_vals)

par(mfrow = c(2, 2))
for (i in 1:4) {
  se <- smooth_estimates(m, smooth = paste0("s(", x_vars[i], ")"), n = 200)
  x_grid <- se[[x_vars[i]]]

  # True function on grid, centered
  f_grid <- true_fns[[i]](x_grid)
  f_grid <- f_grid - mean(f_grid)

  plot(x_grid, se[[".estimate"]], type = "l", lwd = 2, col = "steelblue",
       ylim = range(c(se[[".estimate"]] - 2 * se[[".se"]], se[[".estimate"]] + 2 * se[[".se"]], f_grid)),
       xlab = x_vars[i], ylab = paste0("f(", x_vars[i], ")"),
       main = paste0("s(", x_vars[i], ")"))
  polygon(c(x_grid, rev(x_grid)),
          c(se[[".estimate"]] - 2 * se[[".se"]], rev(se[[".estimate"]] + 2 * se[[".se"]])),
          col = adjustcolor("steelblue", alpha.f = 0.2), border = NA)
  lines(x_grid, f_grid, lty = 2, lwd = 2, col = "red")
  legend("topright", c("estimate", "truth"), lty = c(1, 2),
         col = c("steelblue", "red"), lwd = 2, cex = 0.8)
}
par(mfrow = c(1, 1))
```

Note that $f_3(x_3) = 0$ (the null smooth), and the model should estimate it as approximately flat with an EDF close to 0–1.

## Concurvity

Concurvity is the smooth analogue of collinearity — it measures how well each smooth can be approximated by the other smooth terms. Values close to 1 indicate potential confounding.

```{r}
conc <- concurvity(m, full = TRUE)
print(round(conc, 3))
```

Since our covariates are independent (uniform draws), concurvity should be low.

Pairwise concurvity:

```{r}
conc_pw <- concurvity(m, full = FALSE)
print(lapply(conc_pw, function(x) round(x, 3)))
```

## Comparing models with different k

Does the basis dimension `k` matter? Let us refit with smaller and larger `k`:

```{r}
for (k_val in c(5, 10, 20, 30)) {
  mk <- gam(y ~ s(x0, k = k_val) + s(x1, k = k_val) +
                s(x2, k = k_val) + s(x3, k = k_val),
            data = df, method = "REML")
  edfs <- round(pen.edf(mk), 2)
  dev <- round(summary(mk)$dev.expl * 100, 1)
  cat(sprintf("k=%d  EDF=[%s]  Dev.Expl=%.1f%%\n",
              k_val, paste(edfs, collapse = ", "), dev))
}
```

As long as `k` is large enough to capture the true complexity, results are similar — the penalty takes care of the rest.

## Summary

In this vignette we:

1. Simulated Gu–Wahba data with four smooth functions of varying complexity
2. Fitted a GAM with four additive smooth terms
3. Examined per-smooth EDF via `overview()`
4. Plotted each smooth estimate against the true function
5. Assessed concurvity between smooth terms
6. Demonstrated robustness to the choice of `k`

The next vignette covers non-Gaussian response distributions.
