---
title: "Non-Gaussian Response Distributions"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

GAMs are not limited to Gaussian responses. By specifying a **family** (distribution) and **link function**, we can model count data, binary outcomes, positive continuous data, and more. The general model is:

$$
g(\mu_i) = \beta_0 + f_1(x_{1i}) + \cdots + f_p(x_{pi}), \quad y_i \sim \text{Family}(\mu_i, \phi)
$$

This vignette demonstrates three common non-Gaussian families:

| Family | Link | Response type | Example |
|--------|------|---------------|---------|
| Poisson | log | Counts | Species counts, event rates |
| Binomial | logit | Binary / proportions | Presence–absence, disease status |
| Gamma | inverse (or log) | Positive continuous | Waiting times, claim sizes |

The second half turns to **disease incidence counts**: an *offset* to model
rates rather than counts, and then **overdispersion** handled with `nb()`,
`quasipoisson()` and `tw()`.

## Setup

```{r}
library(mgcv)
library(gratia)
library(ggplot2)
```

## Poisson GAM — count data

### Loading count data

We load counts generated from a Poisson distribution with a smooth log-rate:

$$
y_i \sim \text{Poisson}(\mu_i), \quad \log(\mu_i) = 1 + 1.5 \sin(2\pi x_i)
$$

```{r}
dat_pois <- read.csv("../data_poisson.csv")
x <- dat_pois$x
y_pois <- dat_pois$y
n <- nrow(dat_pois)
eta <- 1.0 + 1.5 * sin(2 * pi * x)

df_pois <- data.frame(x = x, y = y_pois)
```

### Fitting the model

```{r}
m_pois <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_pois,
              family = poisson(link = "log"), method = "REML")
summary(m_pois)
```

### Smooth estimate

```{r}
se_pois <- smooth_estimates(m_pois, n = 200)

ggplot(se_pois, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x, y = eta - mean(eta)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [log scale]", title = "Poisson GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_pois <- residuals(m_pois, type = "deviance")

plot(fitted(m_pois), resid_pois,
     xlab = "Fitted values", ylab = "Deviance residuals",
     main = "Poisson GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Binomial GAM — binary data

### Loading binary data

We load binary outcomes generated from a logistic model:

$$
y_i \sim \text{Bernoulli}(p_i), \quad \text{logit}(p_i) = -0.5 + 2\sin(2\pi x_i)
$$

```{r}
dat_binom <- read.csv("../data_binomial.csv")
x_bin <- dat_binom$x
y_bin <- dat_binom$y
eta_bin <- -0.5 + 2.0 * sin(2 * pi * x_bin)

df_bin <- data.frame(x = x_bin, y = y_bin)
```

### Fitting the model

```{r}
m_bin <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_bin,
             family = binomial(link = "logit"), method = "REML")
summary(m_bin)
```

### Smooth estimate

```{r}
se_bin <- smooth_estimates(m_bin, n = 200)

ggplot(se_bin, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x_bin, y = eta_bin - mean(eta_bin)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [logit scale]", title = "Binomial GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_bin <- residuals(m_bin, type = "deviance")

plot(fitted(m_bin), resid_bin,
     xlab = "Fitted values (probability)", ylab = "Deviance residuals",
     main = "Binomial GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Gamma GAM — positive continuous data

### Loading positive continuous data

We load positive continuous data generated from a Gamma distribution with a smooth log-mean:

$$
y_i \sim \text{Gamma}(\text{shape}, \text{scale}_i), \quad \log(\mu_i) = 1 + \sin(2\pi x_i)
$$

```{r}
dat_gamma <- read.csv("../data_gamma.csv")
x_gam <- dat_gamma$x
y_gam <- dat_gamma$y
eta_gam <- 1.0 + sin(2 * pi * x_gam)

df_gam <- data.frame(x = x_gam, y = y_gam)
```

### Fitting the model

We use `Gamma` with `log` link (more commonly used in practice than the canonical inverse link):

```{r}
m_gam <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_gam,
             family = Gamma(link = "log"), method = "REML")
summary(m_gam)
```

### Smooth estimate

```{r}
se_gam <- smooth_estimates(m_gam, n = 200)

ggplot(se_gam, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x_gam, y = eta_gam - mean(eta_gam)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [log scale]", title = "Gamma GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_gam <- residuals(m_gam, type = "deviance")

plot(fitted(m_gam), resid_gam,
     xlab = "Fitted values", ylab = "Deviance residuals",
     main = "Gamma GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Disease incidence — offsets and exposure

Epidemiological counts depend on how many people are at risk. To model the
**rate** rather than the count we add an **offset**: a term entering the linear
predictor with its coefficient fixed at 1.

$$
y_i \sim \text{Poisson}(\mu_i), \quad
\log(\mu_i) = \log(E_i) + \beta_0 + f(x_i)
$$

We load 400 simulated district-weeks with lognormal populations and a smooth
log-rate:

$$
\log(\text{pop}_i) \sim N(9,\, 0.7^2), \qquad
\log(\mu_i) = \log(\text{pop}_i) - 6.0 + 0.9\sin(2\pi x_i)
$$

```{r}
dat_inc <- read.csv("../data_incidence.csv")
dat_inc$log_pop <- log(dat_inc$pop)
range(dat_inc$pop)
```

### Fitting with an offset

In mgcv the offset may be given either in the formula as `offset(log(pop))` or,
as here, through the `offset` argument:

```{r}
m_inc <- gam(y ~ s(x, k = 12, bs = "cr"), data = dat_inc,
             family = poisson(link = "log"), method = "REML",
             offset = log_pop)
summary(m_inc)
```

```{r}
cat("intercept =", round(coef(m_inc)[1], 4), " (simulated: -6.0)\n")
cat("Pearson dispersion =",
    round(sum(residuals(m_inc, type = "pearson")^2) / m_inc$df.residual, 3), "\n")
```

### Why not model the rate directly?

Estimating the log-population coefficient instead of fixing it at 1:

```{r}
m_cov <- gam(y ~ log_pop + s(x, k = 12, bs = "cr"), data = dat_inc,
             family = poisson(link = "log"), method = "REML")
cat("log(pop) coefficient =", round(coef(m_cov)["log_pop"], 4), "\n")
```

The estimate sits close to 1, but the offset model imposes that relationship
rather than spending a degree of freedom rediscovering it.

### Recovering the rate curve

```{r}
se_inc <- smooth_estimates(m_inc, n = 200)
xg <- se_inc$x
truth_inc <- 0.9 * sin(2 * pi * xg)

ggplot(se_inc, aes(x = x, y = .estimate - mean(.estimate))) +
  geom_ribbon(aes(ymin = .estimate - mean(.estimate) - 2 * .se,
                  ymax = .estimate - mean(.estimate) + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = xg, y = truth_inc - mean(truth_inc)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "environmental index x", y = "log incidence rate (centred)",
       title = "Poisson GAM with log-population offset") +
  theme_minimal()
```

### Prediction requires the offset again

`predict.gam` does **not** add the offset for new data — it must be supplied
explicitly on the link scale:

```{r}
nd <- data.frame(x = c(0.25, 0.75))
eta <- predict(m_inc, newdata = nd, type = "link")
pred_10k <- exp(eta + log(10000))

cat("expected weekly cases in a district of 10,000:\n")
cat("  peak season  (x = 0.25):", round(pred_10k[1], 1), "\n")
cat("  trough       (x = 0.75):", round(pred_10k[2], 1), "\n")
cat("  ratio =", round(pred_10k[1] / pred_10k[2], 2),
    " (simulated:", round(exp(1.8), 2), ")\n")
```

## Overdispersion in count data

Weekly counts at a single surveillance site, simulated with genuine
extra-Poisson variation:

$$
y_i \sim \text{NegBin}(\mu_i, \theta), \quad
\log(\mu_i) = 2.5 + 0.9\sin(2\pi x_i), \quad \theta = 2
$$

```{r}
dat_od <- read.csv("../data_incidence_od.csv")
cat("mean =", round(mean(dat_od$y), 2),
    " variance =", round(var(dat_od$y), 1),
    " zeros =", sum(dat_od$y == 0), "\n")
```

### Detecting overdispersion

```{r}
m_od_pois <- gam(y ~ s(x, k = 12, bs = "cr"), data = dat_od,
                 family = poisson(), method = "REML")
disp_od <- sum(residuals(m_od_pois, type = "pearson")^2) / m_od_pois$df.residual
cat("Pearson dispersion =", round(disp_od, 2), "\n")
```

### Three principled responses

```{r}
m_od_nb <- gam(y ~ s(x, k = 12, bs = "cr"), data = dat_od,
               family = nb(), method = "REML")
cat("nb() theta =", round(m_od_nb$family$getTheta(TRUE), 3), " (simulated: 2.0)\n")
```

```{r}
m_od_qp <- gam(y ~ s(x, k = 12, bs = "cr"), data = dat_od,
               family = quasipoisson(), method = "REML")
cat("quasipoisson scale =", round(m_od_qp$scale, 3), "\n")
```

```{r}
m_od_tw <- gam(y ~ s(x, k = 12, bs = "cr"), data = dat_od,
               family = tw(), method = "REML")
cat("tw() p =", round(m_od_tw$family$getTheta(TRUE), 3),
    " scale =", round(m_od_tw$scale, 3), "\n")
```

```{r}
cat("Poisson AIC =", round(AIC(m_od_pois), 1), "\n")
cat("NegBin  AIC =", round(AIC(m_od_nb), 1), "\n")
```

Quasi-likelihood has no likelihood, so `AIC()` is not meaningful for
`quasipoisson()`.

### What overdispersion costs you

```{r}
se_p  <- smooth_estimates(m_od_pois, n = 50)
se_nb <- smooth_estimates(m_od_nb,   n = 50)
se_qp <- smooth_estimates(m_od_qp,   n = 50)

cat("mean SE of s(x):\n")
cat("  Poisson       ", round(mean(se_p$.se), 4), "\n")
cat("  NegBin        ", round(mean(se_nb$.se), 4),
    " (", round(mean(se_nb$.se) / mean(se_p$.se), 2), "x Poisson)\n")
cat("  quasi-Poisson ", round(mean(se_qp$.se), 4),
    " (", round(mean(se_qp$.se) / mean(se_p$.se), 2), "x Poisson)\n")
```

## Rootograms

The `countreg` package provides `rootogram()`, but it is not on CRAN and may not
be installed; the calculation is short enough to write directly. For each count
$k$ the expected frequency is $\sum_i P(y = k \mid \hat{\mu}_i)$.

```{r}
rootogram_data <- function(counts, mu, family = c("poisson", "nbinom"),
                           theta = NULL, max_count = 40) {
  family <- match.arg(family)
  ks <- 0:max_count
  observed <- as.numeric(table(factor(pmin(counts, max_count), levels = ks)))
  expected <- sapply(ks, function(k) {
    if (family == "poisson") sum(dpois(k, mu)) else sum(dnbinom(k, size = theta, mu = mu))
  })
  data.frame(count = ks, observed = observed, expected = expected,
             sqrt_observed = sqrt(observed), sqrt_expected = sqrt(expected))
}

rg_p  <- rootogram_data(dat_od$y, fitted(m_od_pois), "poisson", max_count = 40)
rg_nb <- rootogram_data(dat_od$y, fitted(m_od_nb), "nbinom",
                        theta = m_od_nb$family$getTheta(TRUE), max_count = 40)
```

In a hanging rootogram each bar hangs from the expected curve by
$\sqrt{\text{observed}}$; bars that miss zero mark counts the model gets wrong.

```{r}
hanging_plot <- function(rg, main) {
  top <- rg$sqrt_expected
  bottom <- rg$sqrt_expected - rg$sqrt_observed
  plot(range(rg$count), range(c(0, top, bottom)), type = "n",
       xlab = "count", ylab = "sqrt(frequency)", main = main)
  rect(rg$count - 0.4, bottom, rg$count + 0.4, top,
       col = adjustcolor("steelblue", 0.6), border = NA)
  lines(rg$count, rg$sqrt_expected, col = "red", lwd = 2)
  abline(h = 0, lty = 2)
}

op <- par(mfrow = c(1, 2))
hanging_plot(rg_p, "Poisson")
hanging_plot(rg_nb, "Negative binomial")
par(op)
```

```{r}
cat("observations equal to 0: ", rg_p$observed[1], "\n")
cat("  expected under Poisson:", round(rg_p$expected[1], 1), "\n")
cat("  expected under NegBin: ", round(rg_nb$expected[1], 1), "\n")
```

The Poisson model expects barely one zero where 15 occur; the negative binomial
expects 16.

## Comparison

```{r}
results <- data.frame(
  Family = c("Poisson", "Binomial", "Gamma"),
  EDF = c(
    round(sum(pen.edf(m_pois)), 2),
    round(sum(pen.edf(m_bin)), 2),
    round(sum(pen.edf(m_gam)), 2)
  ),
  Dev_Explained = c(
    round(summary(m_pois)$dev.expl * 100, 1),
    round(summary(m_bin)$dev.expl * 100, 1),
    round(summary(m_gam)$dev.expl * 100, 1)
  ),
  Scale = c(
    round(summary(m_pois)$scale, 4),
    round(summary(m_bin)$scale, 4),
    round(summary(m_gam)$scale, 4)
  )
)
print(results)
```

## Summary

In this vignette we:

1. Simulated count data and fitted a **Poisson GAM** with a log link
2. Simulated binary data and fitted a **Binomial GAM** with a logit link
3. Simulated positive continuous data and fitted a **Gamma GAM** with a log link
4. Examined smooth estimates and deviance residuals for each family
5. Compared EDF and deviance explained across families
6. Used an **offset** to model incidence rates rather than counts
7. Diagnosed **overdispersion** and handled it with `nb()`, `quasipoisson()`
   and `tw()`
8. Built **rootograms** to see which counts the Poisson model got wrong

Each family uses a different link function to map the linear predictor to the mean of the response distribution, but the smooth estimation machinery is the same.
