---
title: "Non-Gaussian Response Distributions"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

GAMs are not limited to Gaussian responses. By specifying a **family** (distribution) and **link function**, we can model count data, binary outcomes, positive continuous data, and more. The general model is:

$$
g(\mu_i) = \beta_0 + f_1(x_{1i}) + \cdots + f_p(x_{pi}), \quad y_i \sim \text{Family}(\mu_i, \phi)
$$

This vignette demonstrates three common non-Gaussian families:

| Family | Link | Response type | Example |
|--------|------|---------------|---------|
| Poisson | log | Counts | Species counts, event rates |
| Binomial | logit | Binary / proportions | Presence–absence, disease status |
| Gamma | inverse (or log) | Positive continuous | Waiting times, claim sizes |

## Setup

```{r}
library(mgcv)
library(gratia)
library(ggplot2)
```

## Poisson GAM — count data

### Loading count data

We load counts generated from a Poisson distribution with a smooth log-rate:

$$
y_i \sim \text{Poisson}(\mu_i), \quad \log(\mu_i) = 1 + 1.5 \sin(2\pi x_i)
$$

```{r}
dat_pois <- read.csv("../data_poisson.csv")
x <- dat_pois$x
y_pois <- dat_pois$y
n <- nrow(dat_pois)
eta <- 1.0 + 1.5 * sin(2 * pi * x)

df_pois <- data.frame(x = x, y = y_pois)
```

### Fitting the model

```{r}
m_pois <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_pois,
              family = poisson(link = "log"), method = "REML")
summary(m_pois)
```

### Smooth estimate

```{r}
se_pois <- smooth_estimates(m_pois, n = 200)

ggplot(se_pois, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x, y = eta - mean(eta)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [log scale]", title = "Poisson GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_pois <- residuals(m_pois, type = "deviance")

plot(fitted(m_pois), resid_pois,
     xlab = "Fitted values", ylab = "Deviance residuals",
     main = "Poisson GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Binomial GAM — binary data

### Loading binary data

We load binary outcomes generated from a logistic model:

$$
y_i \sim \text{Bernoulli}(p_i), \quad \text{logit}(p_i) = -0.5 + 2\sin(2\pi x_i)
$$

```{r}
dat_binom <- read.csv("../data_binomial.csv")
x_bin <- dat_binom$x
y_bin <- dat_binom$y
eta_bin <- -0.5 + 2.0 * sin(2 * pi * x_bin)

df_bin <- data.frame(x = x_bin, y = y_bin)
```

### Fitting the model

```{r}
m_bin <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_bin,
             family = binomial(link = "logit"), method = "REML")
summary(m_bin)
```

### Smooth estimate

```{r}
se_bin <- smooth_estimates(m_bin, n = 200)

ggplot(se_bin, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x_bin, y = eta_bin - mean(eta_bin)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [logit scale]", title = "Binomial GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_bin <- residuals(m_bin, type = "deviance")

plot(fitted(m_bin), resid_bin,
     xlab = "Fitted values (probability)", ylab = "Deviance residuals",
     main = "Binomial GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Gamma GAM — positive continuous data

### Loading positive continuous data

We load positive continuous data generated from a Gamma distribution with a smooth log-mean:

$$
y_i \sim \text{Gamma}(\text{shape}, \text{scale}_i), \quad \log(\mu_i) = 1 + \sin(2\pi x_i)
$$

```{r}
dat_gamma <- read.csv("../data_gamma.csv")
x_gam <- dat_gamma$x
y_gam <- dat_gamma$y
eta_gam <- 1.0 + sin(2 * pi * x_gam)

df_gam <- data.frame(x = x_gam, y = y_gam)
```

### Fitting the model

We use `Gamma` with `log` link (more commonly used in practice than the canonical inverse link):

```{r}
m_gam <- gam(y ~ s(x, k = 15, bs = "cr"), data = df_gam,
             family = Gamma(link = "log"), method = "REML")
summary(m_gam)
```

### Smooth estimate

```{r}
se_gam <- smooth_estimates(m_gam, n = 200)

ggplot(se_gam, aes(x = x, y = .estimate)) +
  geom_ribbon(aes(ymin = .estimate - 2 * .se, ymax = .estimate + 2 * .se),
              alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_line(data = data.frame(x = x_gam, y = eta_gam - mean(eta_gam)),
            aes(x = x, y = y), linetype = "dashed", linewidth = 1, colour = "red") +
  labs(x = "x", y = "f(x) [log scale]", title = "Gamma GAM — smooth on link scale") +
  theme_minimal()
```

### Deviance residuals

```{r}
resid_gam <- residuals(m_gam, type = "deviance")

plot(fitted(m_gam), resid_gam,
     xlab = "Fitted values", ylab = "Deviance residuals",
     main = "Gamma GAM — residuals", pch = 16, col = adjustcolor("black", 0.4))
abline(h = 0, lty = 2, col = "grey")
```

## Comparison

```{r}
results <- data.frame(
  Family = c("Poisson", "Binomial", "Gamma"),
  EDF = c(
    round(sum(pen.edf(m_pois)), 2),
    round(sum(pen.edf(m_bin)), 2),
    round(sum(pen.edf(m_gam)), 2)
  ),
  Dev_Explained = c(
    round(summary(m_pois)$dev.expl * 100, 1),
    round(summary(m_bin)$dev.expl * 100, 1),
    round(summary(m_gam)$dev.expl * 100, 1)
  ),
  Scale = c(
    round(summary(m_pois)$scale, 4),
    round(summary(m_bin)$scale, 4),
    round(summary(m_gam)$scale, 4)
  )
)
print(results)
```

## Summary

In this vignette we:

1. Simulated count data and fitted a **Poisson GAM** with a log link
2. Simulated binary data and fitted a **Binomial GAM** with a logit link
3. Simulated positive continuous data and fitted a **Gamma GAM** with a log link
4. Examined smooth estimates and deviance residuals for each family
5. Compared EDF and deviance explained across families

Each family uses a different link function to map the linear predictor to the mean of the response distribution, but the smooth estimation machinery is the same.
