---
title: "Extreme Value GAMs"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates extreme value GAMs using the R **evgam** package, fitting the same simulated data as the Julia vignette for comparison.

## Setup

```{r}
library(evgam)
library(mgcv)
```

## GEV model

### Load GEV data

We load the same data as the Julia vignette: block maxima with covariate-dependent location and scale.

```{r}
dat_gev <- read.csv("../data_gev.csv")
n <- nrow(dat_gev)
x <- dat_gev$x
y_gev <- dat_gev$y

mu_true <- 5 + 2 * sin(2 * pi * x)
logsigma_true <- -0.5 + 0.5 * x
sigma_true <- exp(logsigma_true)
xi_true <- 0.1

summary(dat_gev)
```

### Fit the GEV model

In R evgam, we specify a list of formulas—one per distribution parameter.

```{r}
m_gev <- evgam(
  list(y ~ s(x, k = 10, bs = "cr"),   # location μ(x)
       ~ s(x, k = 8, bs = "cr"),       # log-scale ψ(x)
       ~ 1),                            # shape ξ (constant)
  data = dat_gev,
  family = "gev"
)
summary(m_gev)
```

### Examine parameter estimates

```{r}
mu_hat <- predict(m_gev)$location
psi_hat <- predict(m_gev)$logscale
xi_hat <- coef(m_gev)

cat("Location fitted range:", range(mu_hat), "\n")
cat("Location true range:", range(mu_true), "\n")
cat("Correlation (location):", cor(mu_hat, mu_true), "\n")

cat("Log-scale fitted range:", range(psi_hat), "\n")
cat("Log-scale true range:", range(logsigma_true), "\n")
cat("Correlation (log-scale):", cor(psi_hat, logsigma_true), "\n")
```

### Compare fitted vs true

```{r}
rmse_mu <- sqrt(mean((mu_hat - mu_true)^2))
rmse_psi <- sqrt(mean((psi_hat - logsigma_true)^2))
cat("RMSE (location):", round(rmse_mu, 3), "\n")
cat("RMSE (log-scale):", round(rmse_psi, 3), "\n")
```

## GPD model

### Load GPD data

```{r}
dat_gpd <- read.csv("../data_gpd.csv")
n_gpd <- nrow(dat_gpd)
x_gpd <- dat_gpd$x
y_gpd <- dat_gpd$y

logsigma_gpd_true <- 0.5 * sin(2 * pi * x_gpd)
sigma_gpd_true <- exp(logsigma_gpd_true)
xi_gpd_true <- 0.15

summary(dat_gpd)
```

### Fit the GPD model

```{r}
m_gpd <- evgam(
  list(y ~ s(x, k = 10, bs = "cr"),   # log-scale ψ(x)
       ~ 1),                            # shape ξ (constant)
  data = dat_gpd,
  family = "gpd"
)
summary(m_gpd)
```

### Examine GPD estimates

```{r}
psi_gpd_hat <- predict(m_gpd)$logscale
cat("Log-scale fitted range:", range(psi_gpd_hat), "\n")
cat("Log-scale true range:", range(logsigma_gpd_true), "\n")
cat("Correlation (log-scale):", cor(psi_gpd_hat, logsigma_gpd_true), "\n")
```

## Comparison with Julia

The R evgam and Julia GAM.jl `evgam` functions share the same interface structure:

| R evgam | Julia GAM.jl |
|---------|--------------|
| `evgam(list(y ~ s(x), ~ s(x), ~ 1), family="gev")` | `evgam([@gam_formula(y ~ s(x)), @gam_formula(y ~ s(x)), @gam_formula(y ~ 1)], GEVFamily())` |
| `predict(m)$location` | `param_eta(m, 1)` |
| `predict(m)$logscale` | `param_eta(m, 2)` |
| `family = "gev"` | `GEVFamily()` |
| `family = "gpd"` | `GPDFamily()` |
