---
title: "Extreme Value GAMs"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates extreme value GAMs using the R **evgam** package, fitting the same simulated data as the Julia vignette for comparison.

## Setup

```{r}
library(evgam)
library(mgcv)
```

## GEV model

### Load GEV data

We load the same data as the Julia vignette: block maxima with covariate-dependent location and scale.

```{r}
dat_gev <- read.csv("../data_gev.csv")
n <- nrow(dat_gev)
x <- dat_gev$x
y_gev <- dat_gev$y

mu_true <- 5 + 2 * sin(2 * pi * x)
logsigma_true <- -0.5 + 0.5 * x
sigma_true <- exp(logsigma_true)
xi_true <- 0.1

summary(dat_gev)
```

### Fit the GEV model

In R evgam, we specify a list of formulas—one per distribution parameter.

```{r}
m_gev <- evgam(
  list(y ~ s(x, k = 10, bs = "cr"),   # location μ(x)
       ~ s(x, k = 8, bs = "cr"),       # log-scale ψ(x)
       ~ 1),                            # shape ξ (constant)
  data = dat_gev,
  family = "gev"
)
summary(m_gev)
```

### Examine parameter estimates

```{r}
mu_hat <- predict(m_gev)$location
psi_hat <- predict(m_gev)$logscale
xi_hat <- coef(m_gev)

cat("Location fitted range:", range(mu_hat), "\n")
cat("Location true range:", range(mu_true), "\n")
cat("Correlation (location):", cor(mu_hat, mu_true), "\n")

cat("Log-scale fitted range:", range(psi_hat), "\n")
cat("Log-scale true range:", range(logsigma_true), "\n")
cat("Correlation (log-scale):", cor(psi_hat, logsigma_true), "\n")
```

### Compare fitted vs true

```{r}
rmse_mu <- sqrt(mean((mu_hat - mu_true)^2))
rmse_psi <- sqrt(mean((psi_hat - logsigma_true)^2))
cat("RMSE (location):", round(rmse_mu, 3), "\n")
cat("RMSE (log-scale):", round(rmse_psi, 3), "\n")
```

### GEV fitted vs true plots

```{r}
#| fig-width: 8
#| fig-height: 5
ord <- order(x)
x_sorted <- x[ord]

par(mfrow = c(1, 2))

plot(x, y_gev, col = adjustcolor("grey40", alpha.f = 0.3), pch = 16, cex = 0.5,
     xlab = "x", ylab = "Location μ", main = "GEV Location")
lines(x_sorted, mu_hat[ord], col = "steelblue", lwd = 2)
lines(x_sorted, mu_true[ord], col = "red", lty = 2, lwd = 2)
legend("topright", legend = c("Fitted", "True", "Data"),
       col = c("steelblue", "red", adjustcolor("grey40", 0.3)),
       lty = c(1, 2, NA), pch = c(NA, NA, 16), lwd = c(2, 2, NA), cex = 0.8)

plot(x_sorted, psi_hat[ord], type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "Log-scale ψ", main = "GEV Log-scale")
lines(x_sorted, logsigma_true[ord], col = "red", lty = 2, lwd = 2)
legend("topleft", legend = c("Fitted", "True"),
       col = c("steelblue", "red"), lty = c(1, 2), lwd = 2, cex = 0.8)
```

### GEV residual diagnostics

```{r}
#| fig-width: 8
#| fig-height: 5
par(mfrow = c(1, 1))
resid_gev <- y_gev - mu_hat
plot(mu_hat, resid_gev, col = adjustcolor("steelblue", alpha.f = 0.4), pch = 16, cex = 0.5,
     xlab = "Fitted μ", ylab = "Residual (y − μ̂)", main = "GEV Residuals vs Fitted")
abline(h = 0, col = "grey40", lty = 2)
```

## GPD model

### Load GPD data

```{r}
dat_gpd <- read.csv("../data_gpd.csv")
n_gpd <- nrow(dat_gpd)
x_gpd <- dat_gpd$x
y_gpd <- dat_gpd$y

logsigma_gpd_true <- 0.5 * sin(2 * pi * x_gpd)
sigma_gpd_true <- exp(logsigma_gpd_true)
xi_gpd_true <- 0.15

summary(dat_gpd)
```

### Fit the GPD model

```{r}
m_gpd <- evgam(
  list(y ~ s(x, k = 10, bs = "cr"),   # log-scale ψ(x)
       ~ 1),                            # shape ξ (constant)
  data = dat_gpd,
  family = "gpd"
)
summary(m_gpd)
```

### Examine GPD estimates

```{r}
psi_gpd_hat <- predict(m_gpd)$logscale
cat("Log-scale fitted range:", range(psi_gpd_hat), "\n")
cat("Log-scale true range:", range(logsigma_gpd_true), "\n")
cat("Correlation (log-scale):", cor(psi_gpd_hat, logsigma_gpd_true), "\n")
```

### GPD fitted vs true plots

```{r}
#| fig-width: 8
#| fig-height: 5
ord_gpd <- order(x_gpd)
x_gpd_sorted <- x_gpd[ord_gpd]

par(mfrow = c(1, 2))

plot(x_gpd_sorted, psi_gpd_hat[ord_gpd], type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "Log-scale ψ", main = "GPD Log-scale")
lines(x_gpd_sorted, logsigma_gpd_true[ord_gpd], col = "red", lty = 2, lwd = 2)
legend("topright", legend = c("Fitted", "True"),
       col = c("steelblue", "red"), lty = c(1, 2), lwd = 2, cex = 0.8)

sigma_gpd_hat <- exp(psi_gpd_hat)
plot(x_gpd, y_gpd, col = adjustcolor("grey40", alpha.f = 0.3), pch = 16, cex = 0.5,
     xlab = "x", ylab = "y", main = "GPD Data + Fitted Scale")
lines(x_gpd_sorted, sigma_gpd_hat[ord_gpd], col = "steelblue", lwd = 2)
lines(x_gpd_sorted, sigma_gpd_true[ord_gpd], col = "red", lty = 2, lwd = 2)
legend("topright", legend = c("Fitted σ", "True σ", "Data"),
       col = c("steelblue", "red", adjustcolor("grey40", 0.3)),
       lty = c(1, 2, NA), pch = c(NA, NA, 16), lwd = c(2, 2, NA), cex = 0.8)
```

### GPD residual diagnostics

```{r}
#| fig-width: 8
#| fig-height: 5
par(mfrow = c(1, 1))
resid_gpd <- y_gpd / sigma_gpd_hat
plot(sigma_gpd_hat, resid_gpd, col = adjustcolor("steelblue", alpha.f = 0.4), pch = 16, cex = 0.5,
     xlab = "Fitted σ", ylab = "Standardized residual (y / σ̂)",
     main = "GPD Residuals vs Fitted Scale")
abline(h = 1, col = "grey40", lty = 2)
```

## Comparison with Julia

The R evgam and Julia GAM.jl `evgam` functions share the same interface structure:

| R evgam | Julia GAM.jl |
|---------|--------------|
| `evgam(list(y ~ s(x), ~ s(x), ~ 1), family="gev")` | `evgam([@gam_formula(y ~ s(x)), @gam_formula(y ~ s(x)), @gam_formula(y ~ 1)], GEVFamily())` |
| `predict(m)$location` | `param_eta(m, 1)` |
| `predict(m)$logscale` | `param_eta(m, 2)` |
| `family = "gev"` | `GEVFamily()` |
| `family = "gpd"` | `GPDFamily()` |
