---
title: "Seasonal and Group-Varying Smooths"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

This is the mgcv companion to the GAM.jl vignette of the same name, run on the
same simulated CSV data. It covers:

| Construction | Purpose |
|---|---|
| `s(x, bs = "cc")` | cyclic smooth: the curve and its first two derivatives match at the ends |
| `s(x, bs = "cc") + s(t)` | separating a within-year cycle from a multi-year trend |
| `s(x, by = factor)` | one curve per level, **each with its own smoothing parameter** |
| `s(x, by = numeric)` | varying-coefficient term: a slope that changes smoothly with `x` |
| `ti(x, z)` | the *pure interaction*, with both main effects excluded |

## Setup

```{r}
library(mgcv)
library(ggplot2)
```

## The data

```{r}
df_season <- read.csv("../data_season.csv")
df_region <- read.csv("../data_region.csv")
df_region$region <- factor(df_region$region)
cat(sprintf("data_season: %d rows, %d years\n", nrow(df_season), max(df_season$year)))
cat(sprintf("data_region: %d rows, regions = %s\n",
            nrow(df_region), paste(levels(df_region$region), collapse = ", ")))
head(df_season, 4)
```

Both datasets use the same two-harmonic seasonal shape with a period of
exactly 52 weeks:

$$
\text{season}(w) = \sin\!\left(\frac{2\pi w}{52}\right)
                 + 0.35 \sin\!\left(\frac{4\pi w}{52}\right)
$$

```{r}
season <- function(w) sin(2 * pi * w / 52) + 0.35 * sin(4 * pi * w / 52)
cat(sprintf("season(0) = %.3f    season(52) = %.3e\n", season(0), season(52)))
```

`week` runs 0–52 rather than 1–52 so that the observed range equals the period.
mgcv would also let you set the period explicitly with
`knots = list(week = c(0, 52))`; coding the covariate this way gives the same
result and is what the GAM.jl vignette must do, since `gam()` there has no
`knots` argument.

## Cyclic smooths

```{r}
m_cc <- gam(y ~ s(week, k = 12, bs = "cc"), data = df_season, method = "REML")
m_cr <- gam(y ~ s(week, k = 12, bs = "cr"), data = df_season, method = "REML")

ncol_of <- function(m) m$smooth[[1]]$last.para - m$smooth[[1]]$first.para + 1
cat(sprintf("cc: edf = %.3f, %d basis columns\n", sum(pen.edf(m_cc)), ncol_of(m_cc)))
cat(sprintf("cr: edf = %.3f, %d basis columns\n", sum(pen.edf(m_cr)), ncol_of(m_cr)))
```

A cyclic basis with `k = 12` contributes 10 columns against the ordinary cubic
basis's 11, and its null space contains only the constant.

```{r}
ends <- data.frame(week = c(0, 52))
p_cc <- predict(m_cc, ends)
p_cr <- predict(m_cr, ends)
cat(sprintf("cc  f(0) = %.6f   f(52) = %.6f   |difference| = %.3e\n",
            p_cc[1], p_cc[2], abs(p_cc[1] - p_cc[2])))
cat(sprintf("cr  f(0) = %.6f   f(52) = %.6f   |difference| = %.3e\n",
            p_cr[1], p_cr[2], abs(p_cr[1] - p_cr[2])))
```

```{r}
wg <- seq(0, 52, length.out = 300)
grid <- data.frame(week = wg)
plot(wg, predict(m_cc, grid), type = "l", lwd = 2, col = "steelblue",
     xlab = "week of year", ylab = "log abundance",
     main = "Cyclic vs. ordinary cubic basis",
     ylim = range(c(predict(m_cc, grid), predict(m_cr, grid))))
lines(wg, predict(m_cr, grid), lwd = 2, lty = 2, col = "darkorange")
abline(v = c(0, 52), col = "grey", lty = 3)
points(c(0, 52), predict(m_cr, ends), col = "darkorange", pch = 19, cex = 1.2)
legend("bottomright", c("bs = 'cc' (cyclic)", "bs = 'cr'"),
       col = c("steelblue", "darkorange"), lty = c(1, 2), lwd = 2, cex = 0.8)
```

## Separating season from trend

```{r}
m_st <- gam(y ~ s(week, k = 12, bs = "cc") + s(t, k = 10, bs = "cr"),
            data = df_season, method = "REML")
e <- pen.edf(m_st)
for (i in seq_along(m_st$smooth)) {
  cat(sprintf("%-16s edf = %6.3f   sp = %.4f\n",
              m_st$smooth[[i]]$label, e[i], m_st$sp[i]))
}
cat(sprintf("\nDeviance explained: %.1f%%\n", summary(m_st)$dev.expl * 100))
```

Note the trend's smoothing parameter. The trend is nearly a straight line, and
once a term is effectively linear the REML criterion is flat in its smoothing
parameter — mgcv runs it up to a very large value, GAM.jl stops much earlier,
and both report a trend edf of essentially 1. Compare edf, not `sp`.

```{r}
conc <- concurvity(m_st, full = TRUE)
print(round(conc, 4))
```

```{r}
#| fig-width: 10
#| fig-height: 4
par(mfrow = c(1, 2))

se_s <- predict(m_st, data.frame(week = wg, t = mean(df_season$t)),
                type = "terms", se.fit = TRUE)
s_hat <- se_s$fit[, 1] - mean(se_s$fit[, 1])
s_true <- season(wg); s_true <- s_true - mean(s_true)
cat(sprintf("season: correlation with truth = %.5f, RMSE = %.4f\n",
            cor(s_hat, s_true), sqrt(mean((s_hat - s_true)^2))))
plot(wg, s_hat, type = "l", lwd = 2, col = "steelblue",
     xlab = "week", ylab = "seasonal effect", main = "s(week, bs='cc')")
lines(wg, s_true, lty = 2, lwd = 2, col = "red")
legend("bottomright", c("estimate", "truth"), col = c("steelblue", "red"),
       lty = c(1, 2), lwd = 2, cex = 0.8)

tg <- seq(min(df_season$t), max(df_season$t), length.out = 200)
se_t <- predict(m_st, data.frame(week = 0, t = tg), type = "terms")
t_hat <- se_t[, 2] - mean(se_t[, 2])
t_true <- 0.6 * (tg / max(df_season$t))^1.5; t_true <- t_true - mean(t_true)
cat(sprintf("trend : correlation with truth = %.5f, RMSE = %.4f\n",
            cor(t_hat, t_true), sqrt(mean((t_hat - t_true)^2))))
plot(tg, t_hat, type = "l", lwd = 2, col = "steelblue",
     xlab = "t (weeks)", ylab = "trend", main = "s(t, bs='cr')")
lines(tg, t_true, lty = 2, lwd = 2, col = "red")
par(mfrow = c(1, 1))
```

## Factor `by`: a separate curve per group

The three regions differ in both mean level and seasonal amplitude:

| region | level | amplitude |
|---|---|---|
| coastal | 3.4 | 1.40 |
| inland | 3.0 | 0.90 |
| highland | 2.6 | 0.35 |

```{r}
m_by <- gam(y ~ region + s(week, k = 12, bs = "cc", by = region),
            data = df_region, method = "REML")
cat(sprintf("levels    : %s\n", paste(levels(df_region$region), collapse = ", ")))
cat(sprintf("total edf : %.3f over %d coefficients\n",
            sum(pen.edf(m_by)), length(coef(m_by))))
```

The `region` main effect is required: a `by=` smooth is centred within each
level, so it carries each level's shape but not its mean.

Unlike GAM.jl, which reports one combined edf for the whole `by=` term, mgcv
reports one row per level; the two agree when these are summed.

```{r}
amps <- c(coastal = 1.40, highland = 0.35, inland = 0.90)
cat("level      true amplitude    sp        edf\n")
e_by <- pen.edf(m_by)
for (i in seq_along(levels(df_region$region))) {
  lev <- levels(df_region$region)[i]
  cat(sprintf("%-10s %8.2f %14.4f %8.3f\n", lev, amps[[lev]], m_by$sp[i], e_by[i]))
}
```

The smoothing parameter increases as the seasonal signal weakens, exactly as in
GAM.jl. The *values* differ between the packages — mgcv's are roughly an order
of magnitude larger for this basis — but the ordering, the edf and the fitted
values all agree; only the parameterisation of $\lambda$ differs.

```{r}
regions <- levels(df_region$region)
cols <- c("steelblue", "seagreen", "darkorange")
plot(NULL, xlim = c(0, 52), ylim = c(-2, 2), xlab = "week of year",
     ylab = "seasonal effect", main = "s(week, bs='cc', by=region)")
for (i in seq_along(regions)) {
  nd <- data.frame(week = wg, region = factor(regions[i], levels = regions))
  tm <- predict(m_by, nd, type = "terms")
  cn <- paste0("s(week):region", regions[i])
  est <- tm[, cn]; est <- est - mean(est)
  tr <- amps[[regions[i]]] * season(wg); tr <- tr - mean(tr)
  cat(sprintf("%-10s fitted range = %.3f   true range = %.3f   correlation = %.5f\n",
              regions[i], diff(range(est)), diff(range(tr)), cor(est, tr)))
  lines(wg, est, lwd = 2, col = cols[i])
  lines(wg, tr, lwd = 1, lty = 2, col = cols[i])
}
legend("bottomright", regions, col = cols, lwd = 2, cex = 0.8)
```

```{r}
m_shared <- gam(y ~ region + s(week, k = 12, bs = "cc"),
                data = df_region, method = "REML")
cat(sprintf("shared curve   : edf = %6.3f   AIC = %8.2f\n",
            sum(pen.edf(m_shared)), AIC(m_shared)))
cat(sprintf("by = region    : edf = %6.3f   AIC = %8.2f\n",
            sum(pen.edf(m_by)), AIC(m_by)))
cat(sprintf("dAIC = %.2f in favour of per-region curves\n",
            AIC(m_shared) - AIC(m_by)))
```

### `by = factor` or `bs = "fs"`?

- **`s(x, by = f)`** — one smoothing parameter per level; levels treated as
  distinct populations. Requires the factor main effect.
- **`s(x, bs = "fs")`** — a single shared smoothing parameter, levels treated as
  exchangeable, level means absorbed into the smooth. Better when levels are
  many and interchangeable.

## Numeric `by`: varying-coefficient terms

A numeric `by=` multiplies the basis by the covariate, contributing
$z_i f(x_i)$ — a coefficient on $z$ that varies smoothly with $x$. Here the
effect of rainfall varies through the season,
$\beta(w) = 0.45\sin(2\pi(w-8)/52)$.

```{r}
m_vc <- gam(y ~ region + rainfall + s(week, k = 12, bs = "cc", by = region) +
              s(week, k = 12, bs = "cc", by = rainfall),
            data = df_region, method = "REML")
e_vc <- pen.edf(m_vc)
for (i in seq_along(m_vc$smooth)) {
  cat(sprintf("%-28s edf = %.3f\n", m_vc$smooth[[i]]$label, e_vc[i]))
}
```

```{r}
nd <- data.frame(week = wg, region = factor(regions[1], levels = regions),
                 rainfall = 1)
tm <- predict(m_vc, nd, type = "terms")
b_hat <- tm[, "s(week):rainfall"]
b_true <- 0.45 * sin(2 * pi * (wg - 8) / 52)
cat(sprintf("beta(week) recovery: correlation = %.5f, RMSE = %.4f\n",
            cor(b_hat, b_true), sqrt(mean((b_hat - (b_true - mean(b_true)))^2))))
plot(wg, b_hat, type = "l", lwd = 2, col = "steelblue", xlab = "week of year",
     ylab = "effect of rainfall", main = "s(week, bs='cc', by=rainfall)")
lines(wg, b_true - mean(b_true), lty = 2, lwd = 2, col = "red")
abline(h = 0, col = "grey", lty = 3)
legend("bottomright", c("estimate", "truth"), col = c("steelblue", "red"),
       lty = c(1, 2), lwd = 2, cex = 0.8)
```

```{r}
m_const <- gam(y ~ region + rainfall + s(week, k = 12, bs = "cc", by = region),
               data = df_region, method = "REML")
cat(sprintf("constant slope      : AIC = %8.2f\n", AIC(m_const)))
cat(sprintf("varying coefficient : AIC = %8.2f\n", AIC(m_vc)))
cat(sprintf("dAIC = %.2f\n", AIC(m_const) - AIC(m_vc)))
```

## `ti()`: is the seasonality itself changing?

The simulation grew the seasonal amplitude from 0.8 in year 1 to 1.5 in year 8.
`te()` builds a full tensor product that absorbs the marginal main effects;
`ti()` excludes them, leaving the pure interaction, so the main effects can stay
in the model and be tested against.

```{r}
m_main <- gam(y ~ s(week, k = 12, bs = "cc") + s(year, k = 6, bs = "cr"),
              data = df_season, method = "REML")
m_int <- gam(y ~ s(week, k = 12, bs = "cc") + s(year, k = 6, bs = "cr") +
               ti(week, year, bs = c("cc", "cr"), k = 5),
             data = df_season, method = "REML")
cat(sprintf("main effects only : AIC = %8.2f\n", AIC(m_main)))
cat(sprintf("with ti()         : AIC = %8.2f\n", AIC(m_int)))
cat(sprintf("dAIC = %.2f\n", AIC(m_main) - AIC(m_int)))
```

```{r}
summary(m_int)$s.table
```

### `k` in a tensor smooth is per-marginal

`k` sets the basis dimension of each marginal, not the total. In both mgcv and
GAM.jl, `ti(week, year, k = 5)` gives a 12-column block — $3 \times 4$, the
cyclic marginal contributing 3 after its wrap and centring constraints and the
cubic marginal 4 after centring:

```{r}
for (s in m_int$smooth) {
  cat(sprintf("%-18s %2d columns\n", s$label, s$last.para - s$first.para + 1))
}
cat(sprintf("%-18s %2d coefficients\n", "whole model", length(coef(m_int))))
```

```{r}
m_te <- gam(y ~ te(week, year, bs = c("cc", "cr"), k = 5),
            data = df_season, method = "REML")
cat(sprintf("te(week, year, k=5) : %d columns, edf = %.3f, AIC = %.2f\n",
            m_te$smooth[[1]]$last.para - m_te$smooth[[1]]$first.para + 1,
            summary(m_te)$s.table[1, "edf"], AIC(m_te)))
```

Note that `pen.edf()` returns one value per *penalty*, and a tensor product
carries one penalty per marginal — so `sum(pen.edf())` counts a `te` or `ti`
term twice. Use `summary(m)$s.table` for per-smooth degrees of freedom, which
is what GAM.jl's `edf()` reports.

To give the marginals different dimensions, pass a vector: `k = c(8, 4)`.

### What the interaction says

```{r}
f1 <- predict(m_int, data.frame(week = wg, year = 1))
f8 <- predict(m_int, data.frame(week = wg, year = 8))
s_rng <- diff(range(season(wg)))
cat(sprintf("fitted seasonal range: year 1 = %.3f, year 8 = %.3f  (ratio %.2f)\n",
            diff(range(f1)), diff(range(f8)), diff(range(f8)) / diff(range(f1))))
cat(sprintf("true   seasonal range: year 1 = %.3f, year 8 = %.3f  (ratio %.2f)\n",
            0.8 * s_rng, 1.5 * s_rng, 1.5 / 0.8))
plot(wg, f1, type = "l", lwd = 2, col = "steelblue", xlab = "week of year",
     ylab = "fitted log abundance", main = "Seasonal shape by year (ti model)",
     ylim = range(c(f1, f8)))
lines(wg, f8, lwd = 2, col = "darkorange")
legend("bottomright", c("year 1", "year 8"),
       col = c("steelblue", "darkorange"), lwd = 2, cex = 0.8)
```

The interaction is penalized like any other smooth, so shrinkage pulls the two
years' curves towards a common shape and the fitted amplitude ratio is smaller
than the truth — the conservative behaviour wanted from a term whose job is to
answer whether there is anything there at all.

## Summary

In this vignette we:

1. Used a cyclic basis (`bs = "cc"`) to enforce an exact join at the year
   boundary
2. Separated a within-year cycle from a multi-year trend and checked concurvity
3. Fitted a factor `by=` smooth giving each region its own curve and its own
   smoothing parameter
4. Distinguished factor `by=` from `bs = "fs"` and from the numeric `by=`
   varying-coefficient construction
5. Used `ti()` to test whether the seasonal shape changed across years, and
   confirmed that `k` in a tensor smooth is per-marginal
