---
title: "Comparing Smooth Basis Types"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Overview

GAMs represent smooth functions as linear combinations of **basis functions**. The choice of basis affects the shape of the fitted smooth, computational cost, and numerical properties. mgcv supports several basis types:

| Code | Basis | Description |
|------|-------|-------------|
| `"tp"` | Thin plate regression spline | Default. Optimal in a certain sense; no knot placement needed |
| `"cr"` | Cubic regression spline | Cubic spline with knots at data quantiles; efficient for 1D |
| `"ps"` | P-spline | B-spline basis with difference penalty |
| `"gp"` | Gaussian process | Matérn covariance as a basis |
| `"ts"` | Thin plate with shrinkage | As `"tp"`, but the penalty also covers the null space |
| `"cs"` | Cubic with shrinkage | As `"cr"`, but the penalty also covers the null space |
| `"bs"` | B-spline | B-spline basis with an integrated squared-derivative penalty |
| `"cp"` | Cyclic P-spline | P-spline constrained to wrap around |
| `"ad"` | Adaptive smooth | Smoothing parameter varies across the domain |

This vignette fits the same data with each basis and compares the results.
The GAM.jl companion uses the symbols `:tp`, `:cr`, `:ps`, `:gp`, `:ts`,
`:cs`, `:bs`, `:cps` and `:ad` for the same bases — note that mgcv's cyclic
P-spline is `"cp"` where GAM.jl writes `:cps`.

## Setup

```{r}
library(mgcv)
library(gratia)
library(ggplot2)
```

## Loading data

We load $n = 300$ observations generated from a function with both broad and fine-scale structure:

$$
y_i = \sin(2\pi x_i) + 0.5\sin(6\pi x_i) + \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0, 0.5^2)
$$

```{r}
dat <- read.csv("../data.csv")
x <- dat$x
y <- dat$y
n <- nrow(dat)
f_true <- sin(2 * pi * x) + 0.5 * sin(6 * pi * x)
df <- data.frame(x = x, y = y)
```

## Fitting models with different bases

We fit the same formula with each basis type, using `k = 20` basis functions:

```{r}
bases <- c("tp", "cr", "ps", "gp")
models <- list()

for (bs in bases) {
  models[[bs]] <- gam(y ~ s(x, k = 20, bs = bs), data = df, method = "REML")
}
```

## Comparing EDF and deviance

```{r}
results <- data.frame(
  Basis = bases,
  EDF = sapply(models, function(m) round(sum(pen.edf(m)), 2)),
  Deviance = sapply(models, function(m) round(deviance(m), 2)),
  Dev_Explained = sapply(models, function(m) round(summary(m)$dev.expl * 100, 1))
)
print(results)
```

## Comparing smooth estimates

We evaluate each smooth on the same grid and compare:

```{r}
se_list <- lapply(bases, function(bs) {
  se <- smooth_estimates(models[[bs]], n = 200)
  se$basis <- bs
  se
})
se_all <- do.call(rbind, se_list)

ggplot(se_all, aes(x = x, y = .estimate, colour = basis)) +
  geom_line(linewidth = 1) +
  geom_line(data = data.frame(x = x, y = f_true),
            aes(x = x, y = y), inherit.aes = FALSE,
            linetype = "dashed", linewidth = 1) +
  labs(x = "x", y = "f(x)", title = "Smooth estimates by basis type",
       colour = "Basis") +
  theme_minimal()
```

Each basis with its confidence band:

```{r}
#| fig-height: 8
#| fig-width: 8
par(mfrow = c(2, 2))
colors <- c("steelblue", "darkorange", "green4", "purple")

for (i in seq_along(bases)) {
  bs <- bases[i]
  se <- smooth_estimates(models[[bs]], n = 200)
  plot(se$x, se[[".estimate"]], type = "l", lwd = 2, col = colors[i],
       ylim = range(c(se[[".estimate"]] - 2 * se[[".se"]], se[[".estimate"]] + 2 * se[[".se"]])),
       xlab = "x", ylab = "f(x)", main = bs)
  polygon(c(se$x, rev(se$x)),
          c(se[[".estimate"]] - 2 * se[[".se"]], rev(se[[".estimate"]] + 2 * se[[".se"]])),
          col = adjustcolor(colors[i], alpha.f = 0.2), border = NA)
  lines(x, f_true, lty = 2, lwd = 1)
}
par(mfrow = c(1, 1))
```

## Visualizing basis functions

To understand how each basis works, we can examine the model matrix columns for a small number of basis functions (`k = 8`):

```{r}
#| fig-height: 8
#| fig-width: 8
par(mfrow = c(2, 2))
for (i in seq_along(bases)) {
  bs <- bases[i]
  m_small <- gam(y ~ s(x, k = 8, bs = bs), data = df, method = "REML")
  X <- model.matrix(m_small)
  # Remove intercept column
  X_smooth <- predict(m_small, type = "lpmatrix")[, -1]
  k_cols <- ncol(X_smooth)
  ord <- order(df$x)
  matplot(df$x[ord], X_smooth[ord, ], type = "l", lty = 1,
          xlab = "x", ylab = "Basis value",
          main = paste0(bs, " basis (k=8)"))
}
par(mfrow = c(1, 1))
```

## Shrinkage bases: `"ts"` and `"cs"`

An ordinary spline penalty leaves a null space unpenalised, so however large
the smoothing parameter grows a linear trend survives and the term cannot be
removed. The shrinkage bases extend the penalty over the null space, letting a
single smoothing parameter shrink the whole term to zero.

We use a dataset with a relevant covariate `x` and an irrelevant `z`:

```{r}
sh <- read.csv("../data_shrink.csv")

edf_pair <- function(basis, select = FALSE) {
  m <- gam(y ~ s(x, k = 15, bs = basis) + s(z, k = 15, bs = basis),
           data = sh, method = "REML", select = select)
  e <- summary(m)$edf
  c(edf_x = e[1], edf_z = e[2], logsp = log(m$sp))
}

res <- t(sapply(c("tp", "ts", "cr", "cs"), function(b) edf_pair(b)[1:2]))
print(round(res, 4))

sel <- edf_pair("tp", select = TRUE)
cat(sprintf("tp + select: edf(x)=%.3f edf(z)=%.4f\n", sel[1], sel[2]))
```

`"ts"` and `"cs"` both drive the irrelevant term to essentially zero here,
where `"tp"` and `"cr"` leave it around 1.7 effective degrees of freedom.

The smoothing parameters are worth inspecting, because this is where GAM.jl
and mgcv part company:

```{r}
for (b in c("ts", "cs")) {
  m <- gam(y ~ s(x, k = 15, bs = b) + s(z, k = 15, bs = b),
           data = sh, method = "REML")
  cat(sprintf("%s: log(sp) = %s\n", b, paste(round(log(m$sp), 2), collapse = ", ")))
}
```

mgcv drives the irrelevant term's smoothing parameter well past
$\log \lambda = 15$ — for `"cs"` it reaches about 22.6. GAM.jl caps
$\log \lambda$ at 15, which is ample for `:ts` but leaves `:cs` stalled around
`edf(z) ≈ 0.29`. The shrinkage penalties themselves agree; the bound on the
smoothing parameter does not.

## Adaptive smooths: `"ad"`

```{r}
ad <- read.csv("../data_adaptive.csv")
flat <- ad$x < 0.5

for (b in c("ps", "tp", "ad")) {
  m <- gam(y ~ s(x, k = 40, bs = b), data = ad, method = "REML")
  fv <- fitted(m)
  cat(sprintf("%-3s edf=%.2f RMSE(flat)=%.4f RMSE(wiggly)=%.4f\n", b,
      sum(m$edf) - 1,
      sqrt(mean((fv[flat] - ad$f_true[flat])^2)),
      sqrt(mean((fv[!flat] - ad$f_true[!flat])^2))))
}
```

As in GAM.jl, the adaptive basis is roughly three times more accurate on the
flat half while using fewer effective degrees of freedom. `m` sets the number
of adaptive sub-penalties (mgcv's `p.order`), each with its own smoothing
parameter:

```{r}
for (mm in c(3, 5, 8)) {
  m <- gam(y ~ s(x, k = 40, bs = "ad", m = mm), data = ad, method = "REML")
  cat(sprintf("m = %d: %d smoothing parameters, edf = %.2f\n",
              mm, length(m$sp), sum(m$edf) - 1))
}
```

## B-splines and cyclic P-splines

mgcv's `"bs"` takes a **vector** `m = c(spline_order, penalty_order)`. GAM.jl
takes a scalar `m` — the penalty order — and fixes the spline order at `m + 2`,
so GAM.jl's `m = j` corresponds to mgcv's `m = c(3, j)`. The defaults agree.

```{r}
for (mm in list(c(3, 1), c(3, 2), c(3, 3))) {
  m <- gam(y ~ s(x, k = 20, bs = "bs", m = mm), data = df, method = "REML")
  cat(sprintf("m=c(%s): edf=%.3f deviance=%.3f\n",
              paste(mm, collapse = ","), sum(m$edf) - 1, deviance(m)))
}

m_cp <- gam(y ~ s(x, k = 20, bs = "cp"), data = df, method = "REML")
cat(sprintf("cp: edf=%.3f deviance=%.3f\n", sum(m_cp$edf) - 1, deviance(m_cp)))
```

The simulated function is genuinely periodic on $[0, 1]$, so the cyclic
constraint is legitimate and buys a comparable fit for fewer degrees of
freedom.

## Bases without an mgcv counterpart

GAM.jl also provides `:fp` (fractional polynomials) and `:lo` (a loess-style
basis). Neither is an mgcv basis, so there is no side-by-side comparison for
them; they are demonstrated only in the GAM.jl vignette.

## When to use which basis

- **Thin plate (`"tp"`)**: The default choice. Works well in any dimension. Optimal smoothness in a certain mathematical sense. Slightly more expensive than knot-based alternatives for large datasets.

- **Cubic regression spline (`"cr"`)**: Efficient for 1D smoothing with knots at data quantiles. Produces smooth curves that are natural cubic splines. Good default for univariate smooths.

- **P-spline (`"ps"`)**: B-spline basis with a difference penalty on adjacent coefficients. Evenly spaced knots. Computationally efficient and well-behaved, especially for evenly sampled data.

- **Gaussian process (`"gp"`)**: Uses a Matérn covariance kernel by default (the correlation family is chosen through `m`), whose sample paths are finitely differentiable — smoother than an exponential kernel but deliberately rougher than a squared-exponential. A good choice when the underlying function is smooth but not analytically so.

- **Shrinkage (`"ts"`, `"cs"`)**: `"tp"` and `"cr"` with the null space penalised too, so the term can be shrunk out of the model entirely — an alternative to `select = TRUE`.

- **B-spline (`"bs"`)**: A B-spline basis penalising an integrated squared derivative directly, with `m = c(spline_order, penalty_order)`.

- **Cyclic P-spline (`"cp"`)**: For periodic covariates, where the smooth must join up at the ends.

- **Adaptive (`"ad"`)**: For functions whose wiggliness varies across the domain, at the cost of several smoothing parameters instead of one.

## Summary

In this vignette we:

1. Simulated bumpy data with both low- and high-frequency components
2. Fitted GAMs using four different basis types (TP, CR, PS, GP)
3. Compared EDF, deviance, and smooth estimates across bases
4. Visualized the raw basis functions for each type
5. Discussed when each basis is most appropriate

The next vignette demonstrates models with multiple smooth terms.
