---
title: "Shape-Constrained Additive Models"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates shape-constrained additive models using the R **scam** package, fitting the same simulated data as the Julia vignette for comparison.

## Setup

```{r}
library(scam)
library(mgcv)
```

## Example 1: Monotone increasing (dose-response)

### Load data

True function: $f(x) = 3(1 - e^{-5x})$

```{r}
dat <- read.csv("../data.csv")
x <- dat$x
y <- dat$y
n <- nrow(dat)
f_true <- 3.0 * (1.0 - exp(-5.0 * x))
```

### Fit unconstrained GAM vs SCAM

```{r}
m_gam <- gam(y ~ s(x, k = 15, bs = "cr"), data = dat)
m_scam <- scam(y ~ s(x, k = 15, bs = "mpi"), data = dat)
```

### Compare fitted values

```{r}
yhat_gam <- predict(m_gam)
yhat_scam <- predict(m_scam)

rmse_gam <- sqrt(mean((yhat_gam - f_true)^2))
rmse_scam <- sqrt(mean((yhat_scam - f_true)^2))

cat("RMSE (unconstrained GAM):", round(rmse_gam, 4), "\n")
cat("RMSE (SCAM, monotone increasing):", round(rmse_scam, 4), "\n")
```

### Verify monotonicity

```{r}
diffs_scam <- diff(yhat_scam)
diffs_gam <- diff(yhat_gam)

cat("Min successive difference (SCAM):", round(min(diffs_scam), 6), "\n")
cat("All non-decreasing (SCAM):", all(diffs_scam >= -1e-10), "\n")
cat("Min successive difference (GAM):", round(min(diffs_gam), 6), "\n")
cat("All non-decreasing (GAM):", all(diffs_gam >= -1e-10), "\n")
```

### Plot: GAM vs SCAM vs truth

```{r}
#| fig-width: 8
#| fig-height: 5
x_grid <- seq(min(x), max(x), length.out = 200)
nd <- data.frame(x = x_grid)
pred_scam <- predict(m_scam, newdata = nd, se.fit = TRUE)
pred_gam <- predict(m_gam, newdata = nd, se.fit = TRUE)
f_true_grid <- 3 * (1 - exp(-5 * x_grid))

plot(x, y, col = adjustcolor("grey40", 0.4), pch = 16, cex = 0.6,
     xlab = "x", ylab = "f(x)",
     main = "Monotone Increasing: SCAM vs GAM vs Truth")
polygon(c(x_grid, rev(x_grid)),
        c(pred_scam$fit + 2 * pred_scam$se.fit,
          rev(pred_scam$fit - 2 * pred_scam$se.fit)),
        col = adjustcolor("steelblue", 0.2), border = NA)
lines(x_grid, pred_scam$fit, col = "steelblue", lwd = 2)
lines(x_grid, pred_gam$fit, col = "orange", lwd = 2, lty = 3)
lines(x_grid, f_true_grid, col = "red", lwd = 2, lty = 2)
legend("bottomright",
       legend = c("SCAM (mpi) ± 2SE", "GAM (unconstrained)", "True f(x)", "Data"),
       col = c("steelblue", "orange", "red", "grey40"),
       lwd = c(2, 2, 2, NA), lty = c(1, 3, 2, NA),
       pch = c(NA, NA, NA, 16), pt.cex = c(NA, NA, NA, 0.6),
       bg = "white", cex = 0.8)
```

## Example 2: Convex function

### Simulate data

True function: $f(x) = 2x^2$

```{r}
dat_cx <- read.csv("../data_cx.csv")
x_cx <- dat_cx$x
y_cx <- dat_cx$y

dat2 <- data.frame(y = y_cx, x = x_cx)
f_true2 <- 2 * x_cx^2
```

### Fit with convexity constraint

```{r}
m_cx <- scam(y ~ s(x, k = 15, bs = "cx"), data = dat2)

yhat_cx <- predict(m_cx)
rmse_cx <- sqrt(mean((yhat_cx - f_true2)^2))
cat("RMSE (convex SCAM):", round(rmse_cx, 4), "\n")
```

### Verify convexity

```{r}
first_diffs <- diff(yhat_cx)
second_diffs <- diff(first_diffs)
cat("Min second difference:", round(min(second_diffs), 6), "\n")
cat("All convex:", all(second_diffs >= -1e-10), "\n")
```

### Plot: Convex fit and second derivative

```{r}
#| fig-width: 8
#| fig-height: 5
x_grid_cx <- seq(min(x_cx), max(x_cx), length.out = 200)
nd_cx <- data.frame(x = x_grid_cx)
pred_cx <- predict(m_cx, newdata = nd_cx, se.fit = TRUE)
f_true2_grid <- 2 * x_grid_cx^2

par(mfrow = c(1, 2))

# Left: fit vs truth
plot(x_cx, y_cx, col = adjustcolor("grey40", 0.4), pch = 16, cex = 0.6,
     xlab = "x", ylab = "f(x)", main = "Convex Constraint: Fit vs Truth")
polygon(c(x_grid_cx, rev(x_grid_cx)),
        c(pred_cx$fit + 2 * pred_cx$se.fit,
          rev(pred_cx$fit - 2 * pred_cx$se.fit)),
        col = adjustcolor("steelblue", 0.2), border = NA)
lines(x_grid_cx, pred_cx$fit, col = "steelblue", lwd = 2)
lines(x_grid_cx, f_true2_grid, col = "red", lwd = 2, lty = 2)
legend("topleft",
       legend = c("SCAM (cx) ± 2SE", "True f(x) = 2x²", "Data"),
       col = c("steelblue", "red", "grey40"),
       lwd = c(2, 2, NA), lty = c(1, 2, NA),
       pch = c(NA, NA, 16), pt.cex = c(NA, NA, 0.6),
       bg = "white", cex = 0.8)

# Right: numerical second derivative
dx <- diff(x_grid_cx)
first_deriv <- diff(pred_cx$fit) / dx
x_mid <- (x_grid_cx[-1] + x_grid_cx[-length(x_grid_cx)]) / 2
dx2 <- diff(x_mid)
second_deriv <- diff(first_deriv) / dx2
x_mid2 <- (x_mid[-1] + x_mid[-length(x_mid)]) / 2

plot(x_mid2, second_deriv, type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "f''(x)",
     main = "Numerical 2nd Derivative (should be ≥ 0)")
abline(h = 0, col = "red", lwd = 1, lty = 2)

par(mfrow = c(1, 1))
```

## Example 3: Monotone increasing and concave

### Simulate data

True function: $f(x) = 3\sqrt{x}$

```{r}
dat_micv <- read.csv("../data_micv.csv")
x_micv <- dat_micv$x
y_micv <- dat_micv$y

dat3 <- data.frame(y = y_micv, x = x_micv)
f_true3 <- 3 * sqrt(x_micv)
```

### Fit with monotone increasing + concave constraint

```{r}
m_micv <- scam(y ~ s(x, k = 15, bs = "micv"), data = dat3)

yhat_micv <- predict(m_micv)
rmse_micv <- sqrt(mean((yhat_micv - f_true3)^2))
cat("RMSE (monotone increasing + concave):", round(rmse_micv, 4), "\n")
```

### Verify constraints

```{r}
first_diffs_micv <- diff(yhat_micv)
second_diffs_micv <- diff(first_diffs_micv)
cat("Min first difference (monotonicity):", round(min(first_diffs_micv), 6), "\n")
cat("Max second difference (concavity):", round(max(second_diffs_micv), 6), "\n")
cat("Monotone increasing:", all(first_diffs_micv >= -1e-10), "\n")
cat("Concave:", all(second_diffs_micv <= 1e-10), "\n")
```

### Plot: Monotone increasing & concave fit

```{r}
#| fig-width: 8
#| fig-height: 5
x_grid_micv <- seq(min(x_micv), max(x_micv), length.out = 200)
nd_micv <- data.frame(x = x_grid_micv)
pred_micv <- predict(m_micv, newdata = nd_micv, se.fit = TRUE)
f_true3_grid <- 3 * sqrt(x_grid_micv)

plot(x_micv, y_micv, col = adjustcolor("grey40", 0.4), pch = 16, cex = 0.6,
     xlab = "x", ylab = "f(x)",
     main = "Monotone Increasing & Concave: Fit vs Truth")
polygon(c(x_grid_micv, rev(x_grid_micv)),
        c(pred_micv$fit + 2 * pred_micv$se.fit,
          rev(pred_micv$fit - 2 * pred_micv$se.fit)),
        col = adjustcolor("steelblue", 0.2), border = NA)
lines(x_grid_micv, pred_micv$fit, col = "steelblue", lwd = 2)
lines(x_grid_micv, f_true3_grid, col = "red", lwd = 2, lty = 2)
legend("bottomright",
       legend = c("SCAM (micv) ± 2SE", "True f(x) = 3√x", "Data"),
       col = c("steelblue", "red", "grey40"),
       lwd = c(2, 2, NA), lty = c(1, 2, NA),
       pch = c(NA, NA, 16), pt.cex = c(NA, NA, 0.6),
       bg = "white", cex = 0.8)
```

## SCAM model summaries

```{r}
summary(m_scam)
```

```{r}
summary(m_cx)
```

```{r}
summary(m_micv)
```

## Comparison table

| Feature | R `scam` | Julia GAM.jl `scam` |
|---------|----------|---------------------|
| Monotone increasing | `bs="mpi"` | `bs=:mpi` |
| Monotone decreasing | `bs="mpd"` | `bs=:mpd` |
| Convex | `bs="cx"` | `bs=:cx` |
| Concave | `bs="cv"` | `bs=:cv` |
| Mono. inc. + convex | `bs="micx"` | `bs=:micx` |
| Mono. inc. + concave | `bs="micv"` | `bs=:micv` |
| Mono. dec. + convex | `bs="mdcx"` | `bs=:mdcx` |
| Mono. dec. + concave | `bs="mdcv"` | `bs=:mdcv` |
