---
title: "Shape-Constrained Additive Models"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates shape-constrained additive models using the R **scam** package, fitting the same simulated data as the Julia vignette for comparison.

## Setup

```{r}
library(scam)
library(mgcv)
```

## Example 1: Monotone increasing (dose-response)

### Load data

True function: $f(x) = 3(1 - e^{-5x})$

```{r}
dat <- read.csv("../data.csv")
x <- dat$x
y <- dat$y
n <- nrow(dat)
f_true <- 3.0 * (1.0 - exp(-5.0 * x))
```

### Fit unconstrained GAM vs SCAM

```{r}
m_gam <- gam(y ~ s(x, k = 15, bs = "cr"), data = dat)
m_scam <- scam(y ~ s(x, k = 15, bs = "mpi"), data = dat)
```

### Compare fitted values

```{r}
yhat_gam <- predict(m_gam)
yhat_scam <- predict(m_scam)

rmse_gam <- sqrt(mean((yhat_gam - f_true)^2))
rmse_scam <- sqrt(mean((yhat_scam - f_true)^2))

cat("RMSE (unconstrained GAM):", round(rmse_gam, 4), "\n")
cat("RMSE (SCAM, monotone increasing):", round(rmse_scam, 4), "\n")
```

### Verify monotonicity

```{r}
diffs_scam <- diff(yhat_scam)
diffs_gam <- diff(yhat_gam)

cat("Min successive difference (SCAM):", round(min(diffs_scam), 6), "\n")
cat("All non-decreasing (SCAM):", all(diffs_scam >= -1e-10), "\n")
cat("Min successive difference (GAM):", round(min(diffs_gam), 6), "\n")
cat("All non-decreasing (GAM):", all(diffs_gam >= -1e-10), "\n")
```

## Example 2: Convex function

### Simulate data

True function: $f(x) = 2x^2$

```{r}
dat_cx <- read.csv("../data_cx.csv")
x_cx <- dat_cx$x
y_cx <- dat_cx$y

dat2 <- data.frame(y = y_cx, x = x_cx)
f_true2 <- 2 * x_cx^2
```

### Fit with convexity constraint

```{r}
m_cx <- scam(y ~ s(x, k = 15, bs = "cx"), data = dat2)

yhat_cx <- predict(m_cx)
rmse_cx <- sqrt(mean((yhat_cx - f_true2)^2))
cat("RMSE (convex SCAM):", round(rmse_cx, 4), "\n")
```

### Verify convexity

```{r}
first_diffs <- diff(yhat_cx)
second_diffs <- diff(first_diffs)
cat("Min second difference:", round(min(second_diffs), 6), "\n")
cat("All convex:", all(second_diffs >= -1e-10), "\n")
```

## Example 3: Monotone increasing and concave

### Simulate data

True function: $f(x) = 3\sqrt{x}$

```{r}
dat_micv <- read.csv("../data_micv.csv")
x_micv <- dat_micv$x
y_micv <- dat_micv$y

dat3 <- data.frame(y = y_micv, x = x_micv)
f_true3 <- 3 * sqrt(x_micv)
```

### Fit with monotone increasing + concave constraint

```{r}
m_micv <- scam(y ~ s(x, k = 15, bs = "micv"), data = dat3)

yhat_micv <- predict(m_micv)
rmse_micv <- sqrt(mean((yhat_micv - f_true3)^2))
cat("RMSE (monotone increasing + concave):", round(rmse_micv, 4), "\n")
```

### Verify constraints

```{r}
first_diffs_micv <- diff(yhat_micv)
second_diffs_micv <- diff(first_diffs_micv)
cat("Min first difference (monotonicity):", round(min(first_diffs_micv), 6), "\n")
cat("Max second difference (concavity):", round(max(second_diffs_micv), 6), "\n")
cat("Monotone increasing:", all(first_diffs_micv >= -1e-10), "\n")
cat("Concave:", all(second_diffs_micv <= 1e-10), "\n")
```

## SCAM model summaries

```{r}
summary(m_scam)
```

```{r}
summary(m_cx)
```

```{r}
summary(m_micv)
```

## Comparison table

| Feature | R `scam` | Julia GAM.jl `scam` |
|---------|----------|---------------------|
| Monotone increasing | `bs="mpi"` | `bs=:mpi` |
| Monotone decreasing | `bs="mpd"` | `bs=:mpd` |
| Convex | `bs="cx"` | `bs=:cx` |
| Concave | `bs="cv"` | `bs=:cv` |
| Mono. inc. + convex | `bs="micx"` | `bs=:micx` |
| Mono. inc. + concave | `bs="micv"` | `bs=:micv` |
| Mono. dec. + convex | `bs="mdcx"` | `bs=:mdcx` |
| Mono. dec. + concave | `bs="mdcv"` | `bs=:mdcv` |
