---
title: "Migrating from mgcv (R side)"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This companion fits the **same** model on the **same** CSV as the Julia
vignette, so the two printouts can be compared line by line. The quantities
below are the ones the automated parity suite asserts elementwise.

```{r}
#| message: false
library(mgcv)
```

## The model

$y = \sin(x) + \varepsilon$, $\varepsilon \sim N(0, 0.3^2)$, $n = 200$.

```{r}
d <- read.csv("../data.csv")
m <- gam(y ~ s(x, k = 15, bs = "cr"), data = d, method = "REML")
summary(m)
```

## The quantities compared against GAM.jl

```{r}
cat(sprintf("edf   = %.4f\n", sum(summary(m)$edf)))
cat(sprintf("sp    = %.6f  (log scale: %.6f)\n", m$sp[1], log(m$sp[1])))
cat(sprintf("scale = %.6f\n", m$scale))
cat(sprintf("AIC   = %.4f\n", AIC(m)))
```

Prediction standard errors (first five), which GAM.jl matches to a maximum
relative difference of 1.1e-5:

```{r}
p <- predict(m, se.fit = TRUE)
round(head(p$se.fit, 5), 8)
```

The smooth-term table — note the test statistic, which is the one inference
quantity that differs between the packages (GAM.jl uses a documented
simplification of Wood (2013)'s `testStat`; the edf and the p-value conclusion
agree):

```{r}
summary(m)$s.table
```

```{r}
#| fig-width: 7
#| fig-height: 4.5
plot(m, shade = TRUE, residuals = TRUE, pch = 16, cex = 0.4,
     main = "mgcv fit")
```

The Julia vignette's corresponding output is directly comparable: on this
model the two implementations agree on the log smoothing parameter to 9.6e-5,
on the coefficients to 3.7e-6, on prediction standard errors to 1.1e-5, and on
`AIC` to 1.1e-4 (`102.4210` against `102.4209`).
