---
title: "Bayesian GAMs: R Comparison (brms + mgcv)"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This vignette compares Bayesian GAM posterior distributions from **GAM.jl** (Turing.jl MCMC) with two R approaches:

1. **brms** — Full MCMC via Stan's NUTS sampler. brms uses `mgcv::smoothCon()` + `smooth2random()` internally, making it the most direct R analog of GAM.jl's Turing backend. Both use the same smooth2random decomposition with half-normal/exponential priors on smoothing SDs.
2. **mgcv Bayesian posterior approximation** — `mgcv::gam()` computes a Bayesian posterior covariance matrix `Vp` alongside frequentist estimates. We sample from this multivariate normal posterior.

## Setup

```{r message=FALSE}
library(mgcv)
library(brms)
library(MASS)

# Julia posterior samples (from Turing.jl MCMC)
julia_gauss <- read.csv("../posteriors_gaussian_julia.csv")
julia_poisson <- read.csv("../posteriors_poisson_julia.csv")

# brms posterior samples (from Stan MCMC)
brms_gauss <- read.csv("../posteriors_gaussian_brms.csv")
brms_poisson <- read.csv("../posteriors_poisson_brms.csv")

cat(sprintf("Posterior samples: Julia=%d, brms=%d\n",
            nrow(julia_gauss), nrow(brms_gauss)))
```

## Example 1: Gaussian GAM

### Frequentist reference (mgcv)

```{r}
dat <- read.csv("../data_bayes_gaussian.csv")
m <- gam(y ~ s(x, k = 10), data = dat, method = "REML")
summary(m)
```

### Gaussian fit with credible intervals (mgcv Bayesian posterior)

mgcv's `predict()` with `se.fit=TRUE` uses the Bayesian posterior covariance $\mathbf{V}_p$ to produce credible intervals:

```{r fig.width=9, fig.height=6}
# Predict on a fine grid
x_grid <- data.frame(x = seq(min(dat$x), max(dat$x), length.out = 200))
pred <- predict(m, newdata = x_grid, se.fit = TRUE)

# 95% and 80% credible intervals
ci95_lo <- pred$fit - 1.96 * pred$se.fit
ci95_hi <- pred$fit + 1.96 * pred$se.fit
ci80_lo <- pred$fit - 1.282 * pred$se.fit
ci80_hi <- pred$fit + 1.282 * pred$se.fit

plot(dat$x, dat$y, pch = 16, col = adjustcolor("gray50", 0.3), cex = 0.8,
     xlab = "x", ylab = "y",
     main = "Gaussian GAM: data + mgcv Bayesian fit with credible intervals")
polygon(c(x_grid$x, rev(x_grid$x)), c(ci95_lo, rev(ci95_hi)),
        col = adjustcolor("steelblue", 0.15), border = NA)
polygon(c(x_grid$x, rev(x_grid$x)), c(ci80_lo, rev(ci80_hi)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(x_grid$x, pred$fit, col = "steelblue", lwd = 2.5)
legend("topright", c("mgcv Bayesian mean", "80% CI", "95% CI", "data"),
       col = c("steelblue", adjustcolor("steelblue", 0.25),
               adjustcolor("steelblue", 0.15), "gray50"),
       lwd = c(2.5, 8, 8, NA), pch = c(NA, NA, NA, 16), cex = 0.8)
```

### brms fit

The brms model uses the same priors as GAM.jl: `Exponential(1)` on smooth SDs, `Normal(0, 10)` on fixed effects, and `truncated Normal(0, 2.5)` on σ:

```{r eval=FALSE}
m_brms <- brm(
  y ~ s(x, k = 10),
  data = dat,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = "b"),
    prior(exponential(1), class = "sds"),
    prior(normal(0, 2.5), class = "sigma", lb = 0)
  ),
  chains = 2, iter = 2000, warmup = 1000, seed = 2024,
  backend = "cmdstanr"
)
```

### mgcv Bayesian posterior approximation

`mgcv::gam()` computes the Bayesian posterior covariance `Vp` (accounting for smoothing parameter uncertainty). We sample from the multivariate normal posterior $\boldsymbol{\beta} \sim N(\hat{\boldsymbol{\beta}}, \mathbf{V}_p)$ and use a scaled inverse-$\chi^2$ posterior for $\sigma^2$:

```{r}
set.seed(2024)
n_samples <- 4000

# Coefficient posterior
beta_post <- mvrnorm(n_samples, coef(m), vcov(m))

# Residual SD posterior
n <- nrow(dat)
p <- sum(m$edf)
df_resid <- n - p
sigma2_post <- df_resid * m$scale / rchisq(n_samples, df = df_resid)
sigma_post <- sqrt(sigma2_post)
```

### Three-way comparison: GAM.jl vs brms vs mgcv

```{r}
cat("Parameter         | GAM.jl (Turing)   | brms (Stan)       | mgcv (approx)     | Freq\n")
cat("------------------|-------------------|-------------------|-------------------|------\n")
cat(sprintf("sigma  mean       | %.4f             | %.4f             | %.4f             | %.4f\n",
            mean(julia_gauss$sigma_obs), mean(brms_gauss$sigma_obs),
            mean(sigma_post), sqrt(m$scale)))
cat(sprintf("sigma  sd         | %.4f             | %.4f             | %.4f             |\n",
            sd(julia_gauss$sigma_obs), sd(brms_gauss$sigma_obs), sd(sigma_post)))
cat(sprintf("Intercept mean    | %.4f            | %.4f            | %.4f            | %.4f\n",
            mean(julia_gauss$beta_intercept), mean(brms_gauss$beta_intercept),
            mean(beta_post[, 1]), coef(m)[1]))
cat(sprintf("Intercept sd      | %.4f             | %.4f             | %.4f             |\n",
            sd(julia_gauss$beta_intercept), sd(brms_gauss$beta_intercept),
            sd(beta_post[, 1])))
```

### Posterior density plots

```{r fig.width=8, fig.height=5}
par(mfrow = c(1, 2))

# σ_obs posterior histogram
hist(julia_gauss$sigma_obs, breaks = 50, probability = TRUE,
     col = adjustcolor("steelblue", 0.7), border = NA,
     main = expression("Posterior density of " * sigma[obs]),
     xlab = expression(sigma[obs]), ylab = "Density")
abline(v = sqrt(m$scale), col = "red", lwd = 2)
legend("topright", c("GAM.jl posterior", "Frequentist"),
       col = c("steelblue", "red"), lwd = c(10, 2), cex = 0.8)

# Intercept posterior histogram
hist(julia_gauss$beta_intercept, breaks = 50, probability = TRUE,
     col = adjustcolor("steelblue", 0.7), border = NA,
     main = expression("Posterior density of " * beta[1]),
     xlab = expression(beta[1] ~ "(intercept)"), ylab = "Density")
abline(v = coef(m)[1], col = "red", lwd = 2)
legend("topright", c("GAM.jl posterior", "Frequentist"),
       col = c("steelblue", "red"), lwd = c(10, 2), cex = 0.8)
```

### Kolmogorov-Smirnov tests and ECDF correlation

The KS statistic measures the maximum vertical distance between two ECDFs, while the **pairwise ECDF correlation** evaluates both ECDFs on a common grid and computes their Pearson correlation — a value of 1.0 indicates identical distributions:

```{r}
# Helper: pairwise ECDF correlation on a common grid
ecdf_cor <- function(x, y, n_grid = 500) {
  grid <- seq(min(c(x, y)), max(c(x, y)), length.out = n_grid)
  cor(ecdf(x)(grid), ecdf(y)(grid))
}

# GAM.jl vs brms (full MCMC vs full MCMC — should be very similar)
ks_sigma_jb <- ks.test(julia_gauss$sigma_obs, brms_gauss$sigma_obs)
ks_int_jb <- ks.test(julia_gauss$beta_intercept, brms_gauss$beta_intercept)

# GAM.jl vs mgcv (full MCMC vs Gaussian approximation)
ks_sigma_jm <- ks.test(julia_gauss$sigma_obs, sigma_post)
ks_int_jm <- ks.test(julia_gauss$beta_intercept, beta_post[, 1])

# brms vs mgcv
ks_sigma_bm <- ks.test(brms_gauss$sigma_obs, sigma_post)
ks_int_bm <- ks.test(brms_gauss$beta_intercept, beta_post[, 1])

cat("Comparison             | Param     | KS D   | KS p   | ECDF cor\n")
cat("-----------------------|-----------|--------|--------|--------\n")
cat(sprintf("GAM.jl vs brms         | sigma     | %.4f  | %.4f  | %.6f\n",
            ks_sigma_jb$statistic, ks_sigma_jb$p.value,
            ecdf_cor(julia_gauss$sigma_obs, brms_gauss$sigma_obs)))
cat(sprintf("GAM.jl vs brms         | intercept | %.4f  | %.4f  | %.6f\n",
            ks_int_jb$statistic, ks_int_jb$p.value,
            ecdf_cor(julia_gauss$beta_intercept, brms_gauss$beta_intercept)))
cat(sprintf("GAM.jl vs mgcv (approx)| sigma     | %.4f  | %.4f  | %.6f\n",
            ks_sigma_jm$statistic, ks_sigma_jm$p.value,
            ecdf_cor(julia_gauss$sigma_obs, sigma_post)))
cat(sprintf("GAM.jl vs mgcv (approx)| intercept | %.4f  | %.4f  | %.6f\n",
            ks_int_jm$statistic, ks_int_jm$p.value,
            ecdf_cor(julia_gauss$beta_intercept, beta_post[, 1])))
cat(sprintf("brms vs mgcv (approx)  | sigma     | %.4f  | %.4f  | %.6f\n",
            ks_sigma_bm$statistic, ks_sigma_bm$p.value,
            ecdf_cor(brms_gauss$sigma_obs, sigma_post)))
cat(sprintf("brms vs mgcv (approx)  | intercept | %.4f  | %.4f  | %.6f\n",
            ks_int_bm$statistic, ks_int_bm$p.value,
            ecdf_cor(brms_gauss$beta_intercept, beta_post[, 1])))
cat("\n(ECDF correlation near 1.0 = nearly identical distributions)\n")
```

### ECDF comparison plots

```{r fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

# sigma ECDF — three posteriors
plot(ecdf(julia_gauss$sigma_obs), col = "red", main = expression(sigma[obs] ~ "posterior ECDF"),
     xlab = expression(sigma), ylab = "ECDF", lwd = 2)
lines(ecdf(brms_gauss$sigma_obs), col = "darkgreen", lwd = 2)
lines(ecdf(sigma_post), col = "blue", lwd = 2, lty = 2)
abline(v = sqrt(m$scale), lty = 3, col = "black", lwd = 1.5)
legend("bottomright", c("GAM.jl (Turing)", "brms (Stan)", "mgcv (approx)", "Frequentist"),
       col = c("red", "darkgreen", "blue", "black"), lty = c(1, 1, 2, 3), lwd = 2, cex = 0.7)

# Intercept ECDF
plot(ecdf(julia_gauss$beta_intercept), col = "red", main = "Intercept posterior ECDF",
     xlab = expression(beta[0]), ylab = "ECDF", lwd = 2)
lines(ecdf(brms_gauss$beta_intercept), col = "darkgreen", lwd = 2)
lines(ecdf(beta_post[, 1]), col = "blue", lwd = 2, lty = 2)
abline(v = coef(m)[1], lty = 3, col = "black", lwd = 1.5)
legend("bottomright", c("GAM.jl (Turing)", "brms (Stan)", "mgcv (approx)", "Frequentist"),
       col = c("red", "darkgreen", "blue", "black"), lty = c(1, 1, 2, 3), lwd = 2, cex = 0.7)
```

### Q-Q plots: GAM.jl vs brms

The most stringent comparison — both use full MCMC with the same model specification:

```{r fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
probs <- seq(0.01, 0.99, by = 0.01)

# sigma Q-Q
q_julia <- quantile(julia_gauss$sigma_obs, probs)
q_brms <- quantile(brms_gauss$sigma_obs, probs)
plot(q_brms, q_julia, pch = 16, cex = 0.6,
     main = expression(sigma[obs] ~ "Q-Q: GAM.jl vs brms"),
     xlab = "brms quantiles", ylab = "GAM.jl quantiles")
abline(0, 1, col = "red", lwd = 2)

# Intercept Q-Q
q_julia_int <- quantile(julia_gauss$beta_intercept, probs)
q_brms_int <- quantile(brms_gauss$beta_intercept, probs)
plot(q_brms_int, q_julia_int, pch = 16, cex = 0.6,
     main = "Intercept Q-Q: GAM.jl vs brms",
     xlab = "brms quantiles", ylab = "GAM.jl quantiles")
abline(0, 1, col = "red", lwd = 2)
```

## Example 2: Poisson GAM

### Frequentist fit

```{r}
dat2 <- read.csv("../data_bayes_poisson.csv")
m2 <- gam(y ~ s(x, k = 10), data = dat2, family = poisson(), method = "REML")
cat(sprintf("Frequentist intercept (log-scale): %.4f (true: 1.0)\n", coef(m2)[1]))
```

### Poisson fit with credible intervals

```{r fig.width=9, fig.height=6}
x_grid2 <- data.frame(x = seq(min(dat2$x), max(dat2$x), length.out = 200))
pred2 <- predict(m2, newdata = x_grid2, se.fit = TRUE, type = "link")

# On link (log) scale, then transform to response
ci95_lo2 <- exp(pred2$fit - 1.96 * pred2$se.fit)
ci95_hi2 <- exp(pred2$fit + 1.96 * pred2$se.fit)
ci80_lo2 <- exp(pred2$fit - 1.282 * pred2$se.fit)
ci80_hi2 <- exp(pred2$fit + 1.282 * pred2$se.fit)
mu_hat <- exp(pred2$fit)

plot(dat2$x, dat2$y, pch = 16, col = adjustcolor("gray50", 0.3), cex = 0.8,
     xlab = "x", ylab = "count",
     main = "Poisson GAM: data + mgcv Bayesian fit with credible intervals")
polygon(c(x_grid2$x, rev(x_grid2$x)), c(ci95_lo2, rev(ci95_hi2)),
        col = adjustcolor("steelblue", 0.15), border = NA)
polygon(c(x_grid2$x, rev(x_grid2$x)), c(ci80_lo2, rev(ci80_hi2)),
        col = adjustcolor("steelblue", 0.25), border = NA)
lines(x_grid2$x, mu_hat, col = "steelblue", lwd = 2.5)
legend("topright", c("mgcv Bayesian mean", "80% CI", "95% CI", "data"),
       col = c("steelblue", adjustcolor("steelblue", 0.25),
               adjustcolor("steelblue", 0.15), "gray50"),
       lwd = c(2.5, 8, 8, NA), pch = c(NA, NA, NA, 16), cex = 0.8)
```

### mgcv Bayesian posterior

```{r}
beta_post2 <- mvrnorm(n_samples, coef(m2), vcov(m2))
```

### Three-way comparison

```{r}
cat("Parameter         | GAM.jl (Turing)   | brms (Stan)       | mgcv (approx)     | True\n")
cat("------------------|-------------------|-------------------|-------------------|------\n")
cat(sprintf("Intercept mean    | %.4f            | %.4f            | %.4f            | 1.0\n",
            mean(julia_poisson$beta_intercept), mean(brms_poisson$beta_intercept),
            mean(beta_post2[, 1])))
cat(sprintf("Intercept sd      | %.4f             | %.4f             | %.4f             |\n",
            sd(julia_poisson$beta_intercept), sd(brms_poisson$beta_intercept),
            sd(beta_post2[, 1])))
```

### KS tests and ECDF

```{r fig.width=10, fig.height=4}
ks_jb2 <- ks.test(julia_poisson$beta_intercept, brms_poisson$beta_intercept)
ks_jm2 <- ks.test(julia_poisson$beta_intercept, beta_post2[, 1])
ks_bm2 <- ks.test(brms_poisson$beta_intercept, beta_post2[, 1])

cat("Comparison             | KS D   | KS p   | ECDF cor\n")
cat("-----------------------|--------|--------|--------\n")
cat(sprintf("GAM.jl vs brms         | %.4f  | %.4f  | %.6f\n",
            ks_jb2$statistic, ks_jb2$p.value,
            ecdf_cor(julia_poisson$beta_intercept, brms_poisson$beta_intercept)))
cat(sprintf("GAM.jl vs mgcv (approx)| %.4f  | %.4f  | %.6f\n",
            ks_jm2$statistic, ks_jm2$p.value,
            ecdf_cor(julia_poisson$beta_intercept, beta_post2[, 1])))
cat(sprintf("brms vs mgcv (approx)  | %.4f  | %.4f  | %.6f\n",
            ks_bm2$statistic, ks_bm2$p.value,
            ecdf_cor(brms_poisson$beta_intercept, beta_post2[, 1])))

par(mfrow = c(1, 2))

# ECDF
plot(ecdf(julia_poisson$beta_intercept), col = "red",
     main = "Poisson intercept ECDF",
     xlab = expression(beta[0]), ylab = "ECDF", lwd = 2)
lines(ecdf(brms_poisson$beta_intercept), col = "darkgreen", lwd = 2)
lines(ecdf(beta_post2[, 1]), col = "blue", lwd = 2, lty = 2)
abline(v = 1.0, lty = 3, col = "black", lwd = 1.5)
legend("bottomright", c("GAM.jl (Turing)", "brms (Stan)", "mgcv (approx)", "True"),
       col = c("red", "darkgreen", "blue", "black"), lty = c(1, 1, 2, 3), lwd = 2, cex = 0.7)

# Q-Q: GAM.jl vs brms
q_j2 <- quantile(julia_poisson$beta_intercept, probs)
q_b2 <- quantile(brms_poisson$beta_intercept, probs)
plot(q_b2, q_j2, pch = 16, cex = 0.6,
     main = "Poisson intercept Q-Q: GAM.jl vs brms",
     xlab = "brms quantiles", ylab = "GAM.jl quantiles")
abline(0, 1, col = "red", lwd = 2)
```

## Syntax Comparison

| Feature | Julia (GAM.jl + Turing) | R (brms + Stan) | R (mgcv approx) |
|---------|------------------------|-----------------|------------------|
| Bayesian GAM | `gam(f, data; priors=PriorSpec(...))` | `brm(f, data, prior=...)` | `gam(f, data) + mvrnorm` |
| Smooth SD prior | `PriorSpec(sds=Exp(1))` | `prior(exponential(1), class="sds")` | Implicit via REML |
| Residual SD prior | `PriorSpec(sigma=...)` | `prior(..., class="sigma")` | Scaled Inv-χ² |
| Posterior samples | `m.chains[Symbol("σ_obs")]` | `as_draws_df(m)$sigma` | `mvrnorm(n, coef(m), Vp)` |
| MCMC algorithm | NUTS (Turing.jl) | NUTS (Stan) | No MCMC (Gaussian approx) |
| Basis construction | Native Julia | `mgcv::smoothCon()` | Same as brms |

### Key findings

1. **GAM.jl ≈ brms**: Both use full NUTS MCMC with the same smooth2random decomposition and matching priors. KS tests confirm the posteriors are not significantly different — GAM.jl's Turing backend reproduces Stan/brms results.

2. **mgcv approximation is accurate**: The Gaussian posterior approximation from `mgcv::gam()` is close to the full MCMC posteriors for well-identified models, though it cannot capture posterior skewness.

3. **Prior specification**: GAM.jl uses `PriorSpec` with class-level defaults (single call). brms uses a vector of `prior()` statements (one per class). Both are more explicit than mgcv's implicit priors.
