---
title: "Large Data and Spatial Models (R companion)"
author: "Simon Frost"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This companion fits the **same** models on the **same** CSVs as the Julia
vignette, using `mgcv`, so the two printouts can be compared line by line.
All three datasets come from `vignettes/generate_data.jl` with fixed seeds.

```{r}
#| message: false
library(mgcv)
```

## Part 1 — `gam()` versus `bam()`

Same additive model as the Julia side: three `k = 30` cubic-regression
smooths, so $p = 90$ before constraints.

```{r}
dl <- read.csv("../data_large.csv")
f <- y ~ s(x1, k = 30, bs = "cr") + s(x2, k = 30, bs = "cr") + s(x3, k = 30, bs = "cr")

res <- do.call(rbind, lapply(c(1000, 2000, 5000, 10000, 20000), function(n) {
  d <- dl[seq_len(n), ]
  tg <- system.time(mg <- gam(f, data = d, method = "REML"))[["elapsed"]]
  tb <- system.time(mb <- bam(f, data = d, method = "fREML"))[["elapsed"]]
  data.frame(n = n, gam_s = tg, bam_s = tb, speedup = tg / tb,
             edf_gam = sum(mg$edf), edf_bam = sum(mb$edf))
}))
print(res, row.names = FALSE, digits = 4)
```

Note that mgcv's `bam()` wins by a much wider margin than GAM.jl's does. That
is not because GAM.jl's `bam()` is slow — it is because GAM.jl's `gam()` is
already fast, so there is less to recover. Compare the *absolute* times in the
two vignettes rather than the ratios.

`bam(..., discrete = TRUE)` bins covariates onto a grid of unique values and
is faster again; GAM.jl does not implement it, so it is excluded here for a
like-for-like comparison.

## Part 2 — Markov random field

`nb.csv` is a $36 \times 36$ adjacency matrix; mgcv wants a **named list of
neighbour indices**, so we convert. The names must match the factor levels of
`region`.

```{r}
dm <- read.csv("../data_mrf.csv")
dm$region <- factor(dm$region)
A <- as.matrix(read.csv("../nb.csv", check.names = FALSE))
nb <- lapply(seq_len(ncol(A)), function(j) which(A[, j] == 1))
names(nb) <- colnames(A)

mm <- gam(y ~ s(region, bs = "mrf", xt = list(nb = nb)) + s(z, k = 10, bs = "cr"),
          data = dm, method = "REML")
summary(mm)
```

```{r}
truth <- dm$g_true + 1.5 * sin(pi * dm$z)
cat(sprintf("EDF total          = %.3f\n", sum(mm$edf)))
cat(sprintf("scale              = %.4f\n", mm$scale))
cat(sprintf("AIC                = %.3f\n", AIC(mm)))
cat(sprintf("cor(fitted, truth) = %.4f\n", cor(fitted(mm), truth)))
cat(sprintf("sp                 = %s\n", paste(round(mm$sp, 5), collapse = ", ")))
```

The recovered spatial field, on the same lattice layout as the Julia vignette:

```{r}
labels <- levels(dm$region)
grid <- data.frame(region = factor(labels, levels = labels), z = rep(0.5, length(labels)))
ghat <- predict(mm, grid)
ghat <- ghat - mean(ghat)
image(1:6, 1:6, t(matrix(ghat, 6, 6, byrow = TRUE)),
      xlab = "column j", ylab = "row i", main = "MRF estimate (mgcv)",
      col = hcl.colors(32, "viridis"))
```

## Part 3 — Splines on the sphere

`mgcv::s(..., bs = "sos")` takes latitude and longitude in **degrees**, which
is how `data_sphere.csv` stores them. GAM.jl's `bs=:sos` now follows the same
convention (degrees, latitude first), so both fits below read the CSV columns
directly with no conversion.

```{r}
ds <- read.csv("../data_sphere.csv")
m_sos <- gam(y ~ s(lat, lon, k = 50, bs = "sos"), data = ds, method = "REML")
m_tp  <- gam(y ~ s(lat, lon, k = 50, bs = "tp"),  data = ds, method = "REML")

for (nm in c("sos", "tp")) {
  m <- get(paste0("m_", nm))
  cat(sprintf("%-4s EDF=%7.3f  AIC=%9.3f  scale=%.4f  cor=%.4f  RMSE=%.4f\n",
              nm, sum(m$edf), AIC(m), m$scale, cor(fitted(m), ds$f_true),
              sqrt(mean((fitted(m) - ds$f_true)^2))))
}
```

The seam test — two points $2^\circ$ apart on the ground, $358^\circ$ apart in
the `lon` column:

```{r}
seam <- data.frame(lat = c(0, 0), lon = c(179, -179))
ps <- predict(m_sos, seam); pt <- predict(m_tp, seam)
cat(sprintf("sos: %+.4f / %+.4f   |jump| = %.4f\n", ps[1], ps[2], abs(ps[1] - ps[2])))
cat(sprintf("tp:  %+.4f / %+.4f   |jump| = %.4f\n", pt[1], pt[2], abs(pt[1] - pt[2])))
```

```{r}
par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
lat_grid <- seq(-85, 85, length.out = 60)
lon_grid <- seq(-180, 180, length.out = 120)
gp <- expand.grid(lat = lat_grid, lon = lon_grid)
for (nm in c("sos", "tp")) {
  z <- matrix(predict(get(paste0("m_", nm)), gp), nrow = length(lat_grid))
  image(lon_grid, lat_grid, t(z), main = nm, xlab = "longitude", ylab = "latitude",
        col = hcl.colors(32, "Blue-Red 3"))
}
```

## Quantities compared against GAM.jl

The Julia vignette reproduces the tables above on the same CSVs. Both the MRF
and `sos` fits now agree to the printed digits on EDF, AIC, scale and fit
quality: GAM.jl's `bs=:sos` is a direct port of mgcv's spherical-spline
construction, not the geodesic-kernel approximation earlier releases used.
