---
title: "Quantile GAM Regression"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
  pdf:
    toc: true
---

## Introduction

This vignette demonstrates quantile GAM regression using the R **qgam** package, fitting the same heteroscedastic data as the Julia vignette for comparison.

## Setup

```{r}
library(qgam)
library(mgcv)
```

## Load heteroscedastic data

$$y = \sin(2\pi x) + (0.2 + 0.8x)\varepsilon, \quad \varepsilon \sim N(0,1)$$

```{r}
dat <- read.csv("../data.csv")
x <- dat$x
y <- dat$y
n <- nrow(dat)
cat("Data: n =", n, "\n")
cat("y range:", range(y), "\n")
```

```{r}
#| fig-width: 8
#| fig-height: 5
plot(x, y, col = adjustcolor("grey40", alpha.f = 0.3), pch = 16, cex = 0.5,
     xlab = "x", ylab = "y", main = "Heteroscedastic Data: Variance Increases with x")
x_grid <- seq(0, 1, length.out = 200)
lines(x_grid, sin(2 * pi * x_grid), col = "red", lwd = 2)
legend("topleft", legend = c("Data", "True mean: sin(2πx)"),
       col = c("grey40", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))
```

## Fit a single quantile GAM

### Median regression (τ = 0.5)

```{r}
m_50 <- qgam(y ~ s(x, k = 20, bs = "cr"), data = dat, qu = 0.5)
summary(m_50)
```

```{r}
yhat_50 <- predict(m_50)
cat("Median regression fitted range:", range(yhat_50), "\n")
```

### Compare with mean regression

```{r}
m_mean <- gam(y ~ s(x, k = 20, bs = "cr"), data = dat)
yhat_mean <- predict(m_mean)

cat("Mean regression fitted range:", range(yhat_mean), "\n")
cat("Correlation (mean vs median):", cor(yhat_mean, yhat_50), "\n")
```

## Fit multiple quantiles with `mqgam`

```{r}
quantiles <- c(0.1, 0.25, 0.5, 0.75, 0.9)

mq <- mqgam(y ~ s(x, k = 20, bs = "cr"), data = dat, qu = quantiles)
```

Examine fitted ranges for each quantile:

```{r}
for (qu in quantiles) {
  yhat <- qdo(mq, qu, predict, newdata = dat)
  cat(sprintf("τ = %s: fitted range [%.3f, %.3f]\n", qu, min(yhat), max(yhat)))
}
```

```{r}
#| fig-width: 8
#| fig-height: 5
qcolors <- c("0.1" = "red", "0.25" = "orange", "0.5" = "black",
             "0.75" = "dodgerblue", "0.9" = "blue")
plot(x, y, col = adjustcolor("grey40", alpha.f = 0.3), pch = 16, cex = 0.5,
     xlab = "x", ylab = "y", main = "mqgam: Simultaneous Quantile Fits")
for (qu in quantiles) {
  yhat <- qdo(mq, qu, predict, newdata = dat)
  ord_tmp <- order(x)
  lines(x[ord_tmp], yhat[ord_tmp], col = qcolors[as.character(qu)], lwd = 2)
}
legend("topleft", legend = paste("τ =", quantiles), col = qcolors, lwd = 2)
```

## Quantile crossing check

```{r}
ord <- order(x)
predictions <- sapply(quantiles, function(qu) {
  qdo(mq, qu, predict, newdata = dat)[ord]
})

n_crossings <- sum(apply(predictions, 1, is.unsorted))
cat(sprintf("Quantile crossings: %d out of %d observations (%.1f%%)\n",
            n_crossings, n, 100 * n_crossings / n))
```

```{r}
#| fig-width: 8
#| fig-height: 5
qcolors <- c("red", "orange", "black", "dodgerblue", "blue")
matplot(x[ord], predictions, type = "l", lty = 1, lwd = 2, col = qcolors,
        xlab = "x", ylab = "Predicted quantile", main = "Quantile Crossing Check")
crossing_flags <- apply(predictions, 1, is.unsorted)
if (any(crossing_flags)) {
  rug(x[ord][crossing_flags], col = "red", lwd = 2)
}
legend("topleft", legend = paste("τ =", quantiles), col = qcolors, lwd = 2)
mtext(paste0(sum(crossing_flags), " crossings out of ", n, " observations"), side = 3, line = 0.2, cex = 0.8)
```

## Verify quantile coverage

```{r}
for (qu in quantiles) {
  yhat <- qdo(mq, qu, predict, newdata = dat)
  coverage <- mean(y < yhat)
  cat(sprintf("τ = %s: empirical coverage = %.3f (target = %.2f)\n", qu, coverage, qu))
}
```

```{r}
#| fig-width: 8
#| fig-height: 5
coverages <- sapply(quantiles, function(qu) {
  yhat <- qdo(mq, qu, predict, newdata = dat)
  mean(y < yhat)
})
bp <- barplot(coverages, names.arg = paste("τ =", quantiles), col = "steelblue",
              ylim = c(0, 1), ylab = "Coverage",
              main = "Coverage Validation: Empirical vs Target")
points(bp, quantiles, col = "red", pch = 18, cex = 2)
legend("topleft", legend = c("Empirical", "Target"),
       col = c("steelblue", "red"), pch = c(15, 18))
```

## Individual quantile model summaries

```{r}
for (qu in quantiles) {
  cat(sprintf("\n--- Quantile τ = %s ---\n", qu))
  print(summary(qdo(mq, qu)))
}
```

```{r}
#| fig-width: 8
#| fig-height: 5
pinball <- function(y, yhat, qu) {
  r <- y - yhat
  mean(ifelse(r > 0, qu * r, (qu - 1) * r))
}
losses <- sapply(quantiles, function(qu) {
  yhat <- qdo(mq, qu, predict, newdata = dat)
  pinball(y, yhat, qu)
})
barplot(losses, names.arg = paste("τ =", quantiles), col = "steelblue",
        ylab = "Pinball Loss", main = "Pinball Loss by Quantile")
```

## Comparison table

| Feature | R qgam | Julia GAM.jl |
|---------|--------|--------------|
| Single quantile | `qgam(formula, data=dat, qu=τ)` | `qgam(formula, data; qu=τ)` |
| Multiple quantiles | `mqgam(formula, data=dat, qu=c(...))` | `mqgam(formula, data; qu=[...])` |
| Access fits | `mq$fit[["0.5"]]` | `mq.fits[0.5]` |
| Loss function | ELF | ELF |
| Smoothing | Automatic | Automatic |
