---
title: "GAMLSS: R Comparison"
author: "GAM.jl Contributors"
format:
  html:
    code-fold: false
    toc: true
  gfm:
    toc: true
engine: knitr
---

## Overview

This vignette shows the R equivalent of the GAMLSS models fit in `09_gamlss.qmd`, using the `gamlss` package with `gamlss.add` for smooth terms via `ga()`.

## Setup

```{r}
library(gamlss)
library(gamlss.add)
library(mgcv)
```

## Normal Location-Scale

```{r}
df <- read.csv("../data_normal_ls.csv")
cat(sprintf("n = %d, y range: [%.2f, %.2f]\n", nrow(df), min(df$y), max(df$y)))
```

### Fit

```{r}
m <- gamlss(y ~ ga(~s(x, k = 15, bs = "cr")),
            sigma.formula = ~ ga(~s(x, k = 10, bs = "cr")),
            family = NO(), data = df, trace = FALSE,
            control = gamlss.control(n.cyc = 100, trace = FALSE))
summary(m)
```

### Results

```{r}
mu_fit <- fitted(m, "mu")
sigma_fit <- fitted(m, "sigma")
cat(sprintf("μ: cor with truth = %.5f\n", cor(mu_fit, df$mu_true)))
cat(sprintf("σ: cor with truth = %.5f\n", cor(sigma_fit, df$sigma_true)))
```

### Plot

```{r}
par(mfrow = c(1, 2))
plot(df$x, df$y, pch = ".", col = "grey70", main = "Location (μ)",
     xlab = "x", ylab = "y")
lines(df$x, mu_fit, col = "blue", lwd = 2)
lines(df$x, df$mu_true, col = "red", lty = 2, lwd = 2)
legend("topright", c("Fitted", "True"), col = c("blue", "red"), lty = c(1, 2))

plot(df$x, sigma_fit, type = "l", col = "blue", lwd = 2,
     main = "Scale (σ)", xlab = "x", ylab = "σ",
     ylim = range(c(sigma_fit, df$sigma_true)))
lines(df$x, df$sigma_true, col = "red", lty = 2, lwd = 2)
legend("topright", c("Fitted", "True"), col = c("blue", "red"), lty = c(1, 2))
```

## Gamma Location-Scale

```{r}
df_g <- read.csv("../data_gamma_ls.csv")
```

### Fit

The R `gamlss` Gamma family `GA()` uses the same (μ, σ) parameterization as GAM.jl's `GammaLocationScale()`:

```{r}
m_g <- gamlss(y ~ ga(~s(x, k = 15, bs = "cr")),
              sigma.formula = ~ ga(~s(x, k = 10, bs = "cr")),
              family = GA(), data = df_g, trace = FALSE,
              control = gamlss.control(n.cyc = 100, trace = FALSE))
```

### Results

```{r}
mu_g <- fitted(m_g, "mu")
sigma_g <- fitted(m_g, "sigma")
cat(sprintf("μ: cor with truth = %.5f\n", cor(mu_g, df_g$mu_true)))
cat(sprintf("σ: cor with truth = %.5f\n", cor(sigma_g, df_g$sigma_true)))
```

## Beta Regression

```{r}
df_b <- read.csv("../data_beta_reg.csv")
```

### Fit

The R `gamlss` Beta family `BE()` uses (μ, σ) where σ = 1/(1+φ). We use `BEINF0` or `BE` depending on boundary handling:

```{r}
m_b <- gamlss(y ~ ga(~s(x, k = 15, bs = "cr")),
              sigma.formula = ~ 1,
              family = BE(), data = df_b, trace = FALSE,
              control = gamlss.control(n.cyc = 100, trace = FALSE))
```

### Results

```{r}
mu_b <- fitted(m_b, "mu")
cat(sprintf("μ: cor with truth = %.5f\n", cor(mu_b, df_b$mu_true)))
```

## Comparison with mgcv::gaulss

The `mgcv` package also supports some location-scale families via `gaulss()`, `gammals()`, etc.:

```{r}
m_mgcv <- gam(list(y ~ s(x, k = 15, bs = "cr"),
                    ~ s(x, k = 10, bs = "cr")),
              family = gaulss(), data = df)

mu_mgcv <- fitted(m_mgcv)[, 1]
# mgcv gaulss returns 1/σ in the second component
sigma_mgcv <- 1 / fitted(m_mgcv)[, 2]

cat(sprintf("mgcv μ: cor with truth = %.5f\n", cor(mu_mgcv, df$mu_true)))
cat(sprintf("mgcv σ: cor with truth = %.5f\n", cor(sigma_mgcv, df$sigma_true)))
```
